\RequirePackage[svgnames,table]{xcolor}

\documentclass[11pt,letterpaper,logo]{yalearxiv}

\input{preamble}
\input{command}


\title{Kalman Linear Attention}

\runningtitle{Kalman Linear Attention}

\usepackage{fontawesome5}

\definecolor{yaleblue}{RGB}{0,58,112}
\newcommand{\yale}{%
  \textsuperscript{%
    {\usefont{T1}{pbk}{m}{n}\textcolor{yaleblue}{\textbf{Y}}}%
  }%
}

\author{Ngoc Bui\yale, Rex Ying\yale\\
\yale Department of Computer Science, Yale University\\
\faGithub~\textbf{Source Code:} \href{https://github.com/ngocbh/kalman-linear-attention}{\texttt{https://github.com/ngocbh/kalman-linear-attention}}
}

\hypersetup{colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black,
            urlcolor=blue!50!black}

% --- draft-local notation ---
\newcommand{\Fdelta}{\mathsf{F}}               % delta-rule residual block

% Abstract removed: the document class typesets \theabstract inside \maketitle,
% so define it empty (an empty abstract renders as nothing; no heading is added).
\providecommand{\theabstract}{}


\begin{document}


\maketitle


% =============================================================================
\section{Introduction}
\label{sec:intro}

\begin{itemize}[leftmargin=1.4em,itemsep=5pt,topsep=2pt]

  \item \textbf{Fixed-size memory, edited online.} Linear attention replaces the
  growing key--value cache of softmax attention with a fixed-size recurrent state
  $\ve{S}_t \in \R^{d_k \times d_v}$ that is \emph{written} as tokens stream in and
  \emph{read} by the query \citep{katharopoulos2020transformers},
  \begin{equation}
    \text{(write)}\quad \ve{S}_t = \ve{S}_{t-1} + \ve{k}_t \ve{v}_t^{\tr},
    \qquad\qquad
    \text{(read)}\quad \ve{o}_t = \ve{S}_t^{\tr} \ve{q}_t .
    \label{eq:linattn}
  \end{equation}
  The design question is how to \emph{edit} this store online --- what to write and
  what to forget. This plain additive write only ever \emph{adds} associations:
  nothing is removed and non-orthogonal keys interfere, so the store saturates and
  reads degrade.

  \item \textbf{DeltaNet: error-correcting writes.} The delta rule
  \citep{schlag2021linear,yang2024parallelizing} reads what is stored at $\ve{k}_t$ and
  writes back only the correction,
  \begin{equation}
  \begin{aligned}
    \ve{S}_t
      &= \ve{S}_{t-1} + \beta_t \ve{k}_t\,(\ve{v}_t-\ve{S}_{t-1}^{\tr}\ve{k}_t)^{\tr}
        \qquad\text{(canonical residual form)}\\
      &= (\ve{I}-\beta_t \ve{k}_t \ve{k}_t^{\tr})\,\ve{S}_{t-1}
         + \beta_t \ve{k}_t \ve{v}_t^{\tr},
  \end{aligned}
    \label{eq:deltanet}
  \end{equation}
  so a repeated key is \emph{overwritten} rather than accumulated. It still never
  forgets what \emph{other} keys hold.

  \item \textbf{Gated DeltaNet: forgetting by decay.} DeltaNet corrects the current key
  but removes nothing else, so stale associations linger and the state drifts over long
  contexts. To actually \emph{forget}, Gated DeltaNet shrinks the whole state by a
  data-dependent decay $\alpha_t\in(0,1)$ before each write \citep{yang2024gdn},
  \begin{equation}
  \begin{aligned}
    \ve{S}_t
      &= \alpha_t \ve{S}_{t-1}
         + \beta_t \ve{k}_t\,(\ve{v}_t-(\alpha_t \ve{S}_{t-1})^{\tr}\ve{k}_t)^{\tr}
        \qquad\text{(canonical residual form)}\\
      &= \alpha_t\,(\ve{I}-\beta_t \ve{k}_t \ve{k}_t^{\tr})\,\ve{S}_{t-1}
         + \beta_t \ve{k}_t \ve{v}_t^{\tr},
  \end{aligned}
    \label{eq:gdn}
  \end{equation}
  so untouched memories fade geometrically: the state stays bounded and recent tokens
  weigh more. But a single scalar $\alpha_t$ per head fades \emph{every} channel at the
  same rate --- it cannot hold some memories long while dropping others fast.

  \item \textbf{KDA: multi-timescale forgetting.} But a useful memory needs several
  horizons at once --- a name or topic kept for thousands of tokens, a local cue dropped
  immediately. KDA replaces the scalar with a diagonal $\ve{D}_t=\diag(\ve{\alpha}_t)$, giving each
  channel its own decay rate \citep{kimiteam2025kimilinear},
  \begin{equation}
  \begin{aligned}
    \ve{S}_t
      &= \ve{D}_t \ve{S}_{t-1}
         + \beta_t \ve{k}_t\,(\ve{v}_t-(\ve{D}_t \ve{S}_{t-1})^{\tr}\ve{k}_t)^{\tr}
        \qquad\text{(canonical residual form)}\\
      &= (\ve{I}-\beta_t \ve{k}_t \ve{k}_t^{\tr})\,\ve{D}_t\,\ve{S}_{t-1}
         + \beta_t \ve{k}_t \ve{v}_t^{\tr},
  \end{aligned}
    \label{eq:kda}
  \end{equation}
  so one head spans a \emph{mixture of timescales} --- the most expressive forgetting in
  the family, and the baseline we build on.

\end{itemize}


% =============================================================================
\section{Kalman Associative Memory}
\label{sec:memory-kalman}

\subsection{Associative Memory as a State-Space Model}
\label{sec:memory-formulation}

The recurrent state in linear attention is not just a cache compression; it is an
\emph{estimate} of an associative memory. At time $t$, the memory is a matrix
$\ve{S}_t\in\R^{d_k\times d_v}$ that defines the associative map
\begin{equation}
  \mc M_t\colon \R^{d_k}\to\R^{d_v},\qquad \mc M_t(\ve{k})=\ve{S}_t^{\tr}\ve{k} .
  \label{eq:assoc-map}
\end{equation}
The query read is $\ve{o}_t=\mc M_t(\ve{q}_t)=\ve{S}_t^{\tr}\ve{q}_t$. The streaming token provides one
supervised measurement of this map: under key $\ve{k}_t$, the memory should return value $\ve{v}_t$,
i.e.\ $\mc M_t(\ve{k}_t)\approx \ve{v}_t$. This is exactly the shape of a state-space model: an
unobserved state evolves over time, and each token reveals a noisy linear projection of it.

\paragraph{Latent state.}
Let $\ve{S}_t^\star$ be the latent associative map the recurrent state is trying to track. The
transition describes how stored associations persist, decay, or drift before the next
measurement arrives:
\begin{equation}
  \ve{S}_t^\star = \ve{D}_t \ve{S}_{t-1}^\star + \ve{W}_t,\qquad \ve{W}_t\sim\N(0,\ve{Q}_t).
  \label{eq:mem-update-dynamics}
\end{equation}
The operator $\ve{D}_t$ is the process model: it is the formal object corresponding to forgetting
under information drift, topic drift, and other non-stationarity in the stream. For DeltaNet
$\ve{D}_t=\ve{I}$; for Gated DeltaNet $\ve{D}_t=\alpha_t \ve{I}$; for KDA
$\ve{D}_t=\diag(\ve{\alpha}_t)$. The process
noise $\ve{W}_t$ represents memory drift not explained by deterministic decay, while $\ve{Q}_t$ is
its covariance.

\paragraph{Observation.}
The current key reads one direction of the latent map. The observed value is
\begin{equation}
  \ve{v}_t = \ve{S}_t^{\star\tr}\ve{k}_t + \ve{e}_t,\qquad \ve{e}_t\sim\N(0,r_t \ve{I}).
  \label{eq:mem-update-observation}
\end{equation}
Thus $\ve{k}_t^{\tr}$ is the observation map and $\ve{v}_t$ is the measurement. The noise term $\ve{e}_t$
captures the part of the token value that should \emph{not} be treated as a clean memory
target. In an LLM, $\ve{v}_t$ is a contextual feature produced from a noisy token embedding,
local syntax, position, layer mixing, and finite model capacity; not all of it is a stable
fact that can be stored under key $\ve{k}_t$. The covariance $r_t\ve{I}$ sets how reliable this value
observation is: small $r_t$ means trust the token and write strongly, while large $r_t$ means
treat much of $\ve{v}_t$ as context-specific noise.

\paragraph{Filtering problem.}
The state-space model asks for an online filter: given the previous memory estimate and the
new noisy measurement $(\ve{k}_t,\ve{v}_t)$, update the associative memory so it tracks the
non-stationary stream. Forgetting is not an auxiliary gate added to attention; it is part of
the state dynamics that predict how memory changes before the next measurement arrives.


% =============================================================================
\subsection{Kalman Filtering}
\label{sec:kalman-filtering}

The state-space model in \S\ref{sec:memory-formulation} gives a principled solution to the
filtering problem: combine the prior memory estimate with each new measurement so that the
current state estimate has minimal uncertainty. Under the linear--Gaussian assumptions in
\eqref{eq:mem-update-dynamics} and \eqref{eq:mem-update-observation}, this becomes the
Kalman filtering update. The filter carries $\ve{S}_t$ as the current memory estimate, and a
covariance $\ve{P}_t\in\R^{d_k\times d_k}$ tracks uncertainty over key directions in that
estimate.

Tracking $\ve{P}_t$ is necessary because each token observes the associative map only at one key
direction $\ve{k}_t$, and the value $\ve{v}_t$ is noisy. After many tokens, some key directions are
well supported by past measurements, while others remain uncertain or have become stale
because of drift. High uncertainty means the current token should be allowed to edit the
memory strongly; low uncertainty means the prior memory should resist a noisy measurement.

\paragraph{Predict: forgetting before writing.}
Before observing $(\ve{k}_t,\ve{v}_t)$, the filter predicts how the state evolves from step $t-1$ to
step $t$:
\begin{equation}
  \widehat{\ve{S}}_t = \ve{D}_t \ve{S}_{t-1},
  \qquad
  \widehat{\ve{P}}_t = \ve{D}_t\ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t .
  \label{eq:kalman-predict}
\end{equation}
This is the formal role of forgetting. The transition $\ve{D}_t$ moves the associative memory
forward in time, while $\ve{Q}_t$ increases uncertainty for drift that the transition cannot
explain.

\paragraph{Read: compare the prediction to the token.}
The predicted value under the current key is
\begin{equation}
  \widehat{\ve{v}}_t = \widehat{\ve{S}}_t^{\tr}\ve{k}_t,
  \qquad
  \ve{\delta}_t = \ve{v}_t-\widehat{\ve{v}}_t .
  \label{eq:kalman-innovation}
\end{equation}
The vector $\ve{\delta}_t$ is the innovation: the part of the observed value that the predicted
memory does not explain.

\paragraph{Objective: minimize posterior uncertainty.}
After observing $(\ve{k}_t,\ve{v}_t)$, we update the predicted memory state by writing the innovation
back into memory. The only remaining choice is the write direction $\ve{\kappa}_t$: it controls
which key-space direction receives the residual. For a candidate $\ve{\kappa}_t$, the memory
update has the form
\begin{equation}
  \ve{S}_t=\widehat{\ve{S}}_t+\ve{\kappa}_t
  \big(\ve{v}_t-\widehat{\ve{S}}_t^{\tr}\ve{k}_t\big)^{\tr}.
  \label{eq:kalman-residual-write}
\end{equation}
The Kalman objective chooses the write direction that leaves the updated memory estimate as
certain as possible, i.e.\ the one that minimizes posterior uncertainty after the
measurement:
\begin{equation}
  \ve{\kappa}_t^\star
  =
  \arg\min_{\ve{\kappa}_t}
  \Trace\!\left(\ve{P}_t(\ve{\kappa}_t)\right),
  \qquad
  \ve{P}_t(\ve{\kappa}_t)
  =
  (\ve{I}-\ve{\kappa}_t \ve{k}_t^{\tr})\widehat{\ve{P}}_t
  (\ve{I}-\ve{\kappa}_t \ve{k}_t^{\tr})^{\tr}
  + r_t\,\ve{\kappa}_t\ve{\kappa}_t^{\tr}.
  \label{eq:kalman-tracking-objective}
\end{equation}
Here $\ve{P}_t(\ve{\kappa}_t)$ is the posterior covariance that would result if we used the
candidate write direction $\ve{\kappa}_t$. The trace is the total remaining uncertainty across
key-space directions, so minimizing it chooses the residual write that makes the filtered
memory estimate maximally certain after the token is used. The first term in
$\ve{P}_t(\ve{\kappa}_t)$ propagates the predicted uncertainty $\widehat{\ve{P}}_t$ through the correction
operator $\ve{I}-\ve{\kappa}_t\ve{k}_t^{\tr}$: observing key $\ve{k}_t$ should reduce uncertainty in the
directions corrected by the write. The second term, $r_t\ve{\kappa}_t\ve{\kappa}_t^{\tr}$, is the
uncertainty injected by trusting a noisy value measurement. Thus the objective balances two
effects: write strongly where the memory is uncertain, but avoid injecting too much
measurement noise when the token value is unreliable.

\paragraph{Update: write the uncertainty-weighted residual.}
Solving the objective in \eqref{eq:kalman-tracking-objective} gives the Kalman gain, which
decides how much to trust the token relative to the current memory:
\begin{equation}
  \ve{\kappa}_t = \frac{\widehat{\ve{P}}_t \ve{k}_t}
  {r_t+\ve{k}_t^{\tr}\widehat{\ve{P}}_t \ve{k}_t}.
  \label{eq:kalman-gain}
\end{equation}
The optimal filtering update writes only the innovation, along this uncertainty-weighted key
direction:
\begin{equation}
  \ve{S}_t = \widehat{\ve{S}}_t+\ve{\kappa}_t\ve{\delta}_t^{\tr},
  \qquad
  \ve{P}_t=(\ve{I}-\ve{\kappa}_t\ve{k}_t^{\tr})\widehat{\ve{P}}_t .
  \label{eq:kalman-update}
\end{equation}
Finally, the layer reads from the filtered memory,
\begin{equation}
  \ve{o}_t = \ve{S}_t^{\tr}\ve{q}_t .
  \label{eq:kalman-read}
\end{equation}
This filtering update decomposes memory editing into the three operations the model needs:
predict the drifting memory, measure the prediction error at the current key, and write back
the residual in proportion to uncertainty.

\refstepcounter{definition}\label{prop:kalman-optimal-update}
\begin{tcolorbox}[
  title={Proposition~\thedefinition: Kalman Optimal Update},
  colback=blue!7,
  colbacktitle=blue!22,
  colframe=blue!45!black,
  coltitle=black,
  fonttitle=\bfseries,
  boxrule=0.4pt,
  arc=1pt,
  left=5pt,
  right=5pt,
  top=5pt,
  bottom=5pt
]
Under the linear--Gaussian state-space model in
\eqref{eq:mem-update-dynamics}--\eqref{eq:mem-update-observation}, the residual write that
minimizes the posterior uncertainty in \eqref{eq:kalman-tracking-objective} is the Kalman
filtering update below. Given the memory estimate $\ve{S}_{t-1}$ and uncertainty $\ve{P}_{t-1}$
before token $t$, the filter first predicts how the memory drifts under the process model
$\ve{D}_t$ and process noise $\ve{Q}_t$. After observing the key--value pair
$(\ve{k}_t,\ve{v}_t)$, it writes only the value residual
$\ve{v}_t-\widehat{\ve{S}}_t^{\tr}\ve{k}_t$ along the uncertainty-weighted
direction $\ve{\kappa}_t$.
\begin{equation}
\begin{aligned}
  \text{predict:}\quad
  \widehat{\ve{S}}_t &= \ve{D}_t \ve{S}_{t-1},
  &
  \widehat{\ve{P}}_t &= \ve{D}_t \ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t,
  \\[3pt]
  \text{update:}\quad
  \ve{S}_t &= \widehat{\ve{S}}_t+\ve{\kappa}_t
  \big(\ve{v}_t-\widehat{\ve{S}}_t^{\tr}\ve{k}_t\big)^{\tr},
  &
  \ve{P}_t &= (\ve{I}-\ve{\kappa}_t \ve{k}_t^{\tr})\widehat{\ve{P}}_t .
  \\[3pt]
  \text{where}\quad
  \ve{\kappa}_t &= \frac{\widehat{\ve{P}}_t \ve{k}_t}
  {r_t+\ve{k}_t^{\tr}\widehat{\ve{P}}_t \ve{k}_t}.
\end{aligned}
\label{eq:kalman-optimal-update-box}
\end{equation}
\end{tcolorbox}

\subsection{Delta-rule models as fixed-gain Kalman filters}
\label{sec:delta-family-kalman}

The update above separates two modeling choices: the transition $\ve{D}_t$, which predicts how
the memory changes before the token is written, and the gain $\ve{\kappa}_t$, which decides the
key-space direction and strength of the residual write. Delta-rule linear attention is the
fixed-gain special case of this filter. If we discard the covariance recursion and use an
isotropic surrogate $\widehat{\ve{P}}_t\approx b_t \ve{I}$ for the predicted uncertainty, then
\begin{equation}
  \ve{\kappa}_t
  =
  \frac{\widehat{\ve{P}}_t \ve{k}_t}{r_t+\ve{k}_t^{\tr}\widehat{\ve{P}}_t \ve{k}_t}
  \approx
  \frac{b_t}{r_t+b_t\|\ve{k}_t\|_2^2}\,\ve{k}_t
  \equiv
  \beta_t \ve{k}_t .
  \label{eq:fixed-gain-approx}
\end{equation}
For normalized keys, the exact Kalman gain therefore collapses to the scalar write strength
$\beta_t$ used by the delta-rule family. Substituting $\ve{\kappa}_t=\beta_t \ve{k}_t$ into the
Kalman update gives the shared fixed-gain residual form
\begin{equation}
\begin{aligned}
  \ve{S}_t
  &= \ve{D}_t \ve{S}_{t-1}
     + \beta_t \ve{k}_t
     \big(\ve{v}_t-(\ve{D}_t\ve{S}_{t-1})^{\tr}\ve{k}_t\big)^{\tr} \\
  &= (\ve{I}-\beta_t \ve{k}_t \ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}
     + \beta_t \ve{k}_t \ve{v}_t^{\tr}.
\end{aligned}
\label{eq:fixed-gain-delta-family}
\end{equation}
DeltaNet, Gated DeltaNet, and KDA are obtained by choosing different process models:
\begin{equation}
  \text{DeltaNet: }\ve{D}_t=\ve{I},\qquad
  \text{GDN: }\ve{D}_t=\alpha_t \ve{I},\qquad
  \text{KDA: }\ve{D}_t=\diag(\ve{\alpha}_t).
  \label{eq:delta-family-transitions}
\end{equation}
Thus the three models share the same measurement model and residual write; they differ only
in how the memory is predicted before the write. Relative to the Kalman optimal update, they
freeze the uncertainty dynamics and replace the adaptive gain with the fixed first-order
direction $\beta_t \ve{k}_t$. This does not require assuming $\ve{Q}_t=0$; rather, the delta-rule
approximation ignores the covariance prediction
$\widehat{\ve{P}}_t=\ve{D}_t\ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t$ altogether, so the gain no longer depends on the
process model $\ve{D}_t$ or the drift uncertainty $\ve{Q}_t$.


% =============================================================================
\section{Kalman Linear Attention}
\label{sec:kla}

The Kalman optimal update in \eqref{eq:kalman-optimal-update-box} is the target, but the
exact recursion is not directly compatible with fully parallel linear attention. Parallel
training relies on a scan form
\begin{equation}
  \ve{S}_t=\ve{A}_t\ve{S}_{t-1}+\ve{b}_t,
  \label{eq:parallel-affine-scan}
\end{equation}
where the per-token coefficients $\ve{A}_t$ and $\ve{b}_t$ are input-only, or are computable from
associative prefix statistics. Exact Kalman filtering breaks this structure because
$\ve{P}_t$ follows a Riccati recursion and the gain $\ve{\kappa}_t$ depends on the accumulated
posterior uncertainty. Kalman Linear Attention therefore asks which parts of the optimal
update can be relaxed while preserving the scan.

\paragraph{Diagonal information as the tractable state.}
We restrict the uncertainty state to the space of diagonal covariance matrices, the simplest
family that is still more expressive than a scalar gain:
\begin{equation}
  \ve{P}_t\in\{\diag(\ve{p}):\ve{p}\in\R_+^{d_k}\},
  \qquad
  \ve{P}_t=\diag(\ve{p}_t),
  \qquad
  \ve{C}_t=\ve{P}_t^{-1}=\diag(\ve{c}_t),
  \label{eq:diag-cov-info}
\end{equation}
where $\ve{p}_t$ is per-key-direction uncertainty and $\ve{c}_t$ is per-key-direction information.
This is the state-space analogue of diagonal RLS in Appendix~\ref{app:mem-rls}: instead of
learning one scalar write strength, the layer tracks which key channels are already well
measured and which remain uncertain.

\paragraph{Transition-aware diagonal prediction.}
With a diagonal process model $\ve{D}_t=\diag(\ve{\alpha}_t)$ and diagonal process noise
$\ve{Q}_t=\diag(\ve{q}_t)$, the Kalman covariance prediction remains pointwise:
\begin{equation}
  \widehat{\ve{p}}_t = \ve{\alpha}_t^2\odot \ve{p}_{t-1}+\ve{q}_t .
  \label{eq:diag-covariance-prediction}
\end{equation}
The gain is then the diagonal approximation of the Kalman gain,
\begin{equation}
  \ve{\kappa}_t =
  \frac{\widehat{\ve{p}}_t\odot \ve{k}_t}
  {r_t+\sum_i \widehat{\ve{p}}_{t,i}\ve{k}_{t,i}^2}.
  \label{eq:diag-kalman-gain}
\end{equation}
Unlike the fixed-gain delta rule, in the exact prediction~\eqref{eq:diag-covariance-prediction}
the gain $\ve{\kappa}_t$ depends on the same process model that predicts the memory: if a channel
is forgotten or assigned high process noise, its uncertainty changes before the write is formed.
How much of this coupling survives once the recursion is made scan-computable depends on the
parameterization, discussed below.

\paragraph{Parallel gain from M\"obius prefix statistics.}
A direct recurrent implementation would update $\ve{p}_t$ from the posterior covariance at
every token. Kalman Linear Attention instead keeps the information-form sufficient statistics
scan-computable, as in diagonal RLS, while allowing the process noise $\ve{q}_t$ to be
parameterized independently. In information variables,
\eqref{eq:diag-covariance-prediction} becomes
\begin{equation}
  \widehat{\ve{c}}_t
  =
  \frac{\ve{c}_{t-1}}
  {\ve{\alpha}_t^2+\ve{q}_t\odot\ve{c}_{t-1}},
  \qquad
  \widehat{\ve{p}}_t=\frac{1}{\widehat{\ve{c}}_t},
  \label{eq:prefix-diag-uncertainty}
\end{equation}
where products, ratios, and inverses are coordinatewise. The diagonal information update adds
the observation's Fisher information,
\begin{equation}
  \ve{u}_t=\frac{1}{r_t}\,\ve{k}_t\odot\ve{k}_t,
  \qquad
  \ve{c}_t=\widehat{\ve{c}}_t+\ve{u}_t,
  \qquad
  \ve{c}_0=\mu\ve{1}.
  \label{eq:diag-info-update}
\end{equation}
Combining prediction and update gives, for each channel $i$, a linear-fractional recurrence:
\begin{equation}
  c_{t,i}
  =
  \frac{c_{t-1,i}}
  {\alpha_{t,i}^2+q_{t,i}c_{t-1,i}}
  +u_{t,i}
  =
  \frac{(1+u_{t,i}q_{t,i})c_{t-1,i}
        +u_{t,i}\alpha_{t,i}^2}
       {q_{t,i}c_{t-1,i}+\alpha_{t,i}^2}.
  \label{eq:diag-info-mobius}
\end{equation}
Thus each token defines a M\"obius map per channel,
\begin{equation}
  c_{t,i}=f_{t,i}(c_{t-1,i}),
  \qquad
  f_{t,i}(c)=\frac{A_{t,i}c+B_{t,i}}{C_{t,i}c+D_{t,i}},
  \qquad
  \ve{M}_{t,i}
  =
  \begin{bmatrix}
    1+u_{t,i}q_{t,i} & u_{t,i}\alpha_{t,i}^2\\
    q_{t,i} & \alpha_{t,i}^2
  \end{bmatrix}.
  \label{eq:diag-info-mobius-matrix}
\end{equation}
M\"obius maps compose by matrix multiplication, so all prefix information states can be computed
by an associative scan over these $2\times2$ matrices:
\begin{equation}
  \begin{bmatrix}n_{t,i}\\ d_{t,i}\end{bmatrix}
  =
  \ve{M}_{t,i}
  \begin{bmatrix}n_{t-1,i}\\ d_{t-1,i}\end{bmatrix},
  \qquad
  c_{t,i}=\frac{n_{t,i}}{d_{t,i}},
  \qquad
  \begin{bmatrix}n_{0,i}\\ d_{0,i}\end{bmatrix}
  =
  \begin{bmatrix}\mu\\ 1\end{bmatrix}.
  \label{eq:diag-info-scan}
\end{equation}
This keeps the gain transition-aware without imposing the state-proportional restriction
$q_t\propto p_{t-1}$ required by the scalar affine scan.

\paragraph{Parameterization.}
The learnable per-token quantities should be the memory transition and the additive process
noise. Let $\ve{x}_t$ denote the current token representation at this layer. A simple
parameterization is
\begin{equation}
  \begin{aligned}
    \ve{\alpha}_t&=\sigma(\ve{W}_{\alpha}\ve{x}_t+\ve{b}_{\alpha}),
    &
    \ve{q}_t&=q_{\min}+\operatorname{softplus}(\ve{W}_{q}\ve{x}_t+\ve{b}_{q}),
    \\[2pt]
    r_t&=r_{\min}+\operatorname{softplus}(\ve{w}_{r}^{\tr}\ve{x}_t+b_r),
    &
    \ve{D}_t&=\diag(\ve{\alpha}_t).
  \end{aligned}
  \label{eq:kla-parameterization}
\end{equation}
The vector $\ve{\alpha}_t$ parameterizes the deterministic memory transition, while
$\ve{q}_t$ parameterizes absolute process drift in covariance units. In this sense,
$\ve{q}_t$ takes over the learnable role of ``how strongly should this token edit memory?''
from the scalar write gate $\beta_t$ used in GDN and KDA, but indirectly: it changes predicted
uncertainty, and then the gain $\ve{\kappa}_t$ becomes the actual adaptive write direction and
strength. The scalar $r_t$ is the token-dependent observation noise: it controls how reliable
the current value $\ve{v}_t$ is as a memory target. The floors $q_{\min}$, $r_{\min}$, and
$\mu$ can be fixed positive scalars or learned positive scalars per head; $\mu$ is used only to
initialize $\ve{c}_0$, not as a per-token information injection.

\paragraph{Affine memory scan.}
Once $\ve{\kappa}_t$ is computed from prefix statistics, the memory update is again an affine
scan:
\begin{equation}
\begin{aligned}
  \ve{S}_t
  &= \ve{D}_t\ve{S}_{t-1}
     + \ve{\kappa}_t\big(\ve{v}_t-(\ve{D}_t\ve{S}_{t-1})^{\tr}\ve{k}_t\big)^{\tr} \\
  &= (\ve{I}-\ve{\kappa}_t \ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}
     + \ve{\kappa}_t \ve{v}_t^{\tr}.
\end{aligned}
\label{eq:kla-diag-affine-scan}
\end{equation}

\refstepcounter{definition}\label{prop:kalman-linear-attention}
\begin{tcolorbox}[
  title={Proposition~\thedefinition: Kalman Linear Attention},
  colback=blue!7,
  colbacktitle=blue!22,
  colframe=blue!45!black,
  coltitle=black,
  fonttitle=\bfseries,
  boxrule=0.4pt,
  arc=1pt,
  left=5pt,
  right=5pt,
  top=5pt,
  bottom=5pt
]
Kalman Linear Attention is a scan-compatible diagonal-information approximation to the
Kalman optimal update. Given token representation $\ve{x}_t$, key--value pair
$(\ve{k}_t,\ve{v}_t)$, and previous memory $\ve{S}_{t-1}$, the layer parameterizes a
diagonal memory transition, additive process noise, and token-dependent observation noise,
\begin{equation}
\begin{aligned}
  \ve{\alpha}_t &= \sigma(\ve{W}_{\alpha}\ve{x}_t+\ve{b}_{\alpha}),
  &
  \ve{q}_t &= q_{\min}+\operatorname{softplus}(\ve{W}_{q}\ve{x}_t+\ve{b}_{q}),
  &
  r_t &= r_{\min}+\operatorname{softplus}(\ve{w}_{r}^{\tr}\ve{x}_t+b_r),
  \\[2pt]
  \ve{D}_t &= \diag(\ve{\alpha}_t).
\end{aligned}
\label{eq:kla-box-parameterization}
\end{equation}
It then computes transition-aware prefix information statistics by a per-channel M\"obius scan.
With $\ve{u}_t=(\ve{k}_t\odot\ve{k}_t)/r_t$, define for each channel $i$
\begin{equation}
  \ve{M}_{t,i}
  =
  \begin{bmatrix}
    1+u_{t,i}q_{t,i} & u_{t,i}\alpha_{t,i}^2\\
    q_{t,i} & \alpha_{t,i}^2
  \end{bmatrix},
  \qquad
  \begin{bmatrix}n_{t,i}\\ d_{t,i}\end{bmatrix}
  =
  \ve{M}_{t,i}
  \begin{bmatrix}n_{t-1,i}\\ d_{t-1,i}\end{bmatrix},
  \qquad
  c_{t,i}=\frac{n_{t,i}}{d_{t,i}},
  \qquad
  \begin{bmatrix}n_{0,i}\\ d_{0,i}\end{bmatrix}
  =
  \begin{bmatrix}\mu\\ 1\end{bmatrix}.
  \label{eq:kla-box-mobius}
\end{equation}
The adaptive gain at token $t$ uses the exclusive prefix state $\ve{c}_{t-1}$:
\begin{equation}
\begin{aligned}
  \widehat{\ve{c}}_t
  &=
  \frac{\ve{c}_{t-1}}
  {\ve{\alpha}_t^2+\ve{q}_t\odot\ve{c}_{t-1}},
  &
  \widehat{\ve{p}}_t
  &= \frac{1}{\widehat{\ve{c}}_t},
  \\[3pt]
  \ve{\kappa}_t
  &=
  \frac{\widehat{\ve{p}}_t\odot\ve{k}_t}
  {r_t+\sum_i \widehat{\ve{p}}_{t,i}\ve{k}_{t,i}^2}.
\end{aligned}
\label{eq:kla-box-gain}
\end{equation}
With $\ve{\kappa}_t$ available before the memory scan, the memory update is an affine scan:
\begin{equation}
\begin{aligned}
  \ve{S}_t
  &=
  (\ve{I}-\ve{\kappa}_t\ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}
  +\ve{\kappa}_t\ve{v}_t^{\tr},
  &
  \ve{o}_t &= \ve{S}_t^{\tr}\ve{q}_t .
\end{aligned}
\label{eq:kla-box-memory}
\end{equation}
Thus KLA preserves the Kalman structure of predicting memory through $\ve{D}_t$ and writing
an uncertainty-weighted residual, while replacing the dense covariance recursion by
parallel prefix statistics.
\end{tcolorbox}

Thus the diagonal approximation preserves the two ingredients missing from fixed-gain
delta-rule models while keeping the training primitive unchanged: the memory is predicted
through $\ve{D}_t$, and the residual write is weighted by a transition-aware uncertainty
estimate.

\paragraph{Reduction to fixed-gain delta rules.}
The gain \eqref{eq:diag-kalman-gain} collapses to a fixed-direction delta-rule write exactly when
the predicted uncertainty is isotropic, $\widehat{\ve{P}}_t\approx b_t\ve{I}$, equivalently
$\widehat{\ve{p}}_t=b_t\ve{1}$: then
$\ve{\kappa}_t=\frac{b_t}{r_t+b_t\|\ve{k}_t\|_2^2}\ve{k}_t\equiv\beta_t\ve{k}_t$ and
\eqref{eq:kla-diag-affine-scan} becomes
$\ve{S}_t=(\ve{I}-\beta_t\ve{k}_t\ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}+\beta_t\ve{k}_t\ve{v}_t^{\tr}$,
the vanilla delta-rule residual update with per-token write gate $\beta_t$ and transition
$\ve{D}_t=\diag(\ve{\alpha}_t)$. The transition alone then selects the baseline, with the
\emph{same} isotropic gain in every case: $\ve{\alpha}_t=\ve{1}$ gives DeltaNet, a uniform
$\ve{\alpha}_t$ gives GatedDeltaNet, and a \emph{per-channel} $\ve{\alpha}_t$ gives KDA. For KDA,
this means choosing $\ve{q}_t$ so that
$\widehat{\ve{p}}_{t,i}=\alpha_{t,i}^2 p_{t-1,i}+q_{t,i}=b_t$ for every channel, equivalently
$q_{t,i}=b_t-\alpha_{t,i}^2 p_{t-1,i}$ when this is nonnegative. KLA strictly generalizes these
fixed-gain rules by allowing $\widehat{\ve{p}}_t$ to be anisotropic, so the write strength can
vary across key channels with the accumulated evidence.


% =============================================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Synthetic Experiments}
\label{sec:synthetic-experiments}

We evaluate Kalman Linear Attention on the synthetic long-context tests used by
Kimi Linear~\citep{kimiteam2025kimilinear}. These tasks isolate the finite-memory
behaviors that motivate KLA: exact copying, associative recall, and state tracking under
many interleaved updates. Following their setup, all synthetic runs use a small language
model with $2$ layers, $2$ attention heads, and head dimension $128$. For each task, we train
for at most $20{,}000$ steps and select the best training curve over the learning-rate grid
$\{5\times 10^{-5}, 10^{-4}, 5\times 10^{-4}, 10^{-3}\}$. We report two views of the same
protocol: peak accuracy as the sequence length increases from $256$ to $2048$ tokens, and
convergence speed at a fixed context length of $1024$ tokens.

\paragraph{Palindrome.}
The palindrome task tests exact copying from a compressed recurrent memory. The model is
given a random token sequence followed by a separator and must emit the sequence in reverse
order. For example, after observing
\[
  \texttt{O G R S U N E <sep>},
\]
the target continuation is
\[
  \texttt{E N U S R G O}.
\]
This task stresses whether the attention mechanism can preserve token-level information
across the whole prefix rather than only maintaining coarse statistics.

\paragraph{Multi-query associative recall.}
Multi-query associative recall (MQAR) tests whether the model can bind multiple keys to
their values and retrieve several queried values at the end of the context. A typical input
contains interleaved key--value pairs, then a separator, then query keys; for example,
\[
  \texttt{A 1 C 3 B 0 M 8 G 5 E 4 <sep> B G},
\]
with target outputs
\[
  \texttt{0 5}
\]
at the query positions. This task directly probes interference in fixed-size associative
memory: the model must retain the relevant bindings while ignoring unrelated pairs.

\paragraph{Stack state tracking.}
The state-tracking task follows the stack benchmark used in Kimi Linear. The stream contains
operations over $64$ independent last-in-first-out stacks, each identified by an ID. A push
operation such as \texttt{<push> 1 G} adds token \texttt{G} to stack $1$, while a pop query for
stack $0$ asks the model to predict the most recently pushed token on that stack. The model
must therefore maintain many evolving latent states and update only the addressed stack at
each operation. This setting tests whether KLA can selectively overwrite stale information
without erasing unrelated memory channels.


% =============================================================================
\section{Future Directions}
\label{sec:future-directions}

The diagonal-information construction is the conservative path. The same state-space view
also suggests several extensions that trade additional structure for a closer approximation
to the Kalman update.

\paragraph{Direction A: richer process models.}
\begin{enumerate}[leftmargin=1.8em,itemsep=3pt,topsep=2pt,label=\textbf{A\arabic*.}]
  \item \textbf{Block-diagonal transitions.} Use
  $\ve{D}_t=\operatorname{blockdiag}(\ve{D}_t^{(1)},\ldots,\ve{D}_t^{(m)})$. Small blocks allow local
  mixing among related memory channels while keeping the transition structured enough for
  parallel composition.

  \item \textbf{Basis-rotated diagonal transitions.} Use
  $\ve{D}_t=\ve{R}\diag(\ve{\alpha}_t)\ve{R}^{\tr}$ with a fixed or slowly changing basis $\ve{R}$. Forgetting then
  acts in learned memory directions rather than raw coordinates, but the per-token decay
  remains diagonal in that basis.

  \item \textbf{Prefix-conditioned transitions.} Let $\ve{D}_t$ depend on scan-computable
  statistics $\ve{z}_t$, such as decayed key energy or an innovation proxy, instead of depending
  directly on the sequential state $\ve{S}_{t-1}$. Then $\ve{D}_t=\ve{D}(\ve{x}_t,\ve{z}_t)$ can react to drift while
  keeping $\ve{A}_t$ available before the scan.
\end{enumerate}

\paragraph{Direction B: richer covariance and gain families.}
\begin{enumerate}[leftmargin=1.8em,itemsep=3pt,topsep=2pt,label=\textbf{B\arabic*.}]
  \item \textbf{Transition-aware scalar covariance.} Set $\ve{P}_t=c_t\ve{I}$, but predict the scalar
  uncertainty through $\ve{D}_t$ before forming the gain. This still recovers the delta-rule
  direction $\ve{\kappa}_t=\beta_t \ve{k}_t$, while allowing the scalar write strength to depend on
  how strongly the memory has just been forgotten.

  \item \textbf{Block-diagonal covariance.} Track $\ve{P}_t$ or $\ve{C}_t=\ve{P}_t^{-1}$ in small blocks.
  This captures correlations among nearby key channels while replacing one large inverse by
  many small inverses. The prediction remains local:
  $\widehat{\ve{P}}_t^{(j)}=\ve{D}_t^{(j)}\ve{P}_{t-1}^{(j)}\ve{D}_t^{(j)\tr}+\ve{Q}_t^{(j)}$.

  \item \textbf{Diagonal-plus-low-rank covariance.} Use
  $\ve{P}_t=\diag(\ve{p}_t)+\ve{U}_t\ve{U}_t^{\tr}$, with $\ve{p}_t$ and $\ve{U}_t$ produced from prefix statistics.
  Woodbury-style reads can keep the gain tractable, while the low-rank term captures global
  uncertainty directions missed by a diagonal approximation.

  \item \textbf{Chunk-boundary covariance updates.} Update $\ve{P}_t$ only at chunk boundaries,
  while freezing the gain inside each chunk. This gives a TTT-like approximation: the
  expensive uncertainty update is carried by a short sequential pass across chunks, but all
  token updates within a chunk remain fixed-gain affine scans.
\end{enumerate}


\bibliography{main}
\bibliographystyle{plainnat}


\appendix

% =============================================================================
\section{Background: the Kalman filter}
\label{app:kalman}

The Kalman filter is the optimal recursive estimator for a linear--Gaussian
state-space model; \S\ref{sec:memory-kalman} casts the linear-attention memory as one
instance. We recall the classical form and its specializations here.

\paragraph{Linear--Gaussian state-space model.}
A hidden state $\ve{x}_t\in\R^n$ evolves and is observed through
\begin{equation}
\begin{aligned}
  \ve{x}_t &= \ve{F}_t \ve{x}_{t-1} + \ve{w}_t,
  & \ve{w}_t &\sim \N(0,\ve{Q}_t)\quad\text{(process / dynamics)},\\[2pt]
  \ve{y}_t &= \ve{H}_t \ve{x}_t + \ve{e}_t,
  & \ve{e}_t &\sim \N(0,\ve{R}_t)\quad\text{(observation)},
\end{aligned}
\label{eq:kf-ssm}
\end{equation}
with transition $\ve{F}_t$, observation map $\ve{H}_t$, and independent zero-mean Gaussian
process and measurement noise of covariance $\ve{Q}_t$ and $\ve{R}_t$. Given $\ve{y}_{1:t}$ the
posterior over the state is Gaussian, $p(\ve{x}_t\mid \ve{y}_{1:t})=\N(\widehat{\ve{x}}_t,\ve{P}_t)$, and the
filter propagates its mean $\widehat{\ve{x}}_t$ and covariance $\ve{P}_t$.

\paragraph{The filtering recursion.}
Each step is a \emph{predict} (time update) then an \emph{update} (measurement update):
\begin{equation}
\begin{aligned}
  \text{predict:}\quad & \widehat{\ve{x}}_t^- = \ve{F}_t \widehat{\ve{x}}_{t-1},
  & \ve{P}_t^- &= \ve{F}_t \ve{P}_{t-1} \ve{F}_t^{\tr} + \ve{Q}_t,\\[2pt]
  \text{update:}\quad  & \widehat{\ve{x}}_t = \widehat{\ve{x}}_t^- + \ve{K}_t\,\ve{\nu}_t,
  & \ve{P}_t &= (\ve{I} - \ve{K}_t \ve{H}_t)\,\ve{P}_t^-,
\end{aligned}
\label{eq:kf-rec}
\end{equation}
driven by the \emph{innovation} (prediction error) and \emph{Kalman gain}
\begin{equation}
  \ve{\nu}_t = \ve{y}_t - \ve{H}_t \widehat{\ve{x}}_t^-, \qquad
  \ve{K}_t = \ve{P}_t^- \ve{H}_t^{\tr}\,\ve{\Sigma}_t^{-1}, \qquad
  \ve{\Sigma}_t = \ve{H}_t \ve{P}_t^- \ve{H}_t^{\tr} + \ve{R}_t ,
  \label{eq:kf-gain}
\end{equation}
where $\ve{\Sigma}_t$ is the innovation covariance. The gain weighs new evidence against
the prior by relative uncertainty: $\ve{K}_t$ is large when the prior is uncertain ($\ve{P}_t^-$
large) or the measurement reliable ($\ve{R}_t$ small), and small otherwise.

\paragraph{Optimality.}
For a linear--Gaussian model \eqref{eq:kf-ssm} the filter is exact: $\widehat{\ve{x}}_t$ is the
posterior mean, simultaneously the minimum-mean-square-error and maximum-a-posteriori
estimate. Without Gaussianity it remains the best \emph{linear} unbiased estimator. It
is thus the optimal way to fuse a drifting prior with streaming observations.

\paragraph{Special case: recursive least squares.}
Take a \emph{static} state ($\ve{F}_t=\ve{I}$, $\ve{Q}_t=0$) observed through a regressor,
$y_t=\ve{\phi}_t^{\tr}\ve{\theta}+e_t$ (so $\ve{H}_t=\ve{\phi}_t^{\tr}$, scalar $R_t=r$). Then
\eqref{eq:kf-rec}--\eqref{eq:kf-gain} reduce to \emph{recursive least squares}, which
minimizes $\sum_{\tau\le t}\|y_\tau-\ve{\phi}_\tau^{\tr}\ve{\theta}\|_2^2$ online; the covariance
$\ve{P}_t=\big(\sum_{\tau\le t}\ve{\phi}_\tau\ve{\phi}_\tau^{\tr}\big)^{-1}$ is the inverse data Gram
matrix. An exponential \emph{forgetting factor} $\lambda\in(0,1]$ --- weighting old
data by $\lambda^{t-\tau}$ --- is exactly the Kalman filter with process noise
$\ve{Q}_t=(\lambda^{-1}-1)\ve{P}_{t-1}$ (equivalently covariance inflation $\ve{P}_t^-=\ve{P}_{t-1}/\lambda$),
letting the estimate track slowly drifting parameters. RLS is therefore the
static-state (or random-walk) special case of the Kalman filter.

% =============================================================================
\subsection{The associative-memory Kalman filter}
\label{app:mem-kf}

\paragraph{State-space model.}
Write the memory itself as the hidden state: it drifts between tokens and is observed, one
noisy key--value pair at a time, through the current key,
\begin{equation}
\begin{aligned}
  \ve{S}_t &= \ve{D}_t \ve{S}_{t-1} + \ve{W}_t, & \ve{W}_t &\sim\N(0,\ve{Q}_t),\\[2pt]
  \ve{v}_t &= \ve{S}_t^{\tr}\ve{k}_t + \ve{e}_t, & \ve{e}_t &\sim\N(0,r_t),
\end{aligned}
\label{eq:mem-ssm}
\end{equation}
with transition/decay $\ve{D}_t\in\R^{d_k\times d_k}$, drift covariance $\ve{Q}_t$, and scalar
observation noise $r_t$. The observation map $\ve{k}_t^{\tr}$ and the dynamics $\ve{D}_t$ are shared by
all $d_v$ value channels, so the posterior factorizes across them: we track one estimate
$\ve{S}_t\in\R^{d_k\times d_v}$ and a single key-space covariance $\ve{P}_t\in\R^{d_k\times d_k}$.

\paragraph{Optimal solution.}
The minimum-variance estimate is the Kalman recursion \eqref{eq:kf-rec}--\eqref{eq:kf-gain}
specialized to \eqref{eq:mem-ssm}:
\begin{equation}
\begin{aligned}
  \text{predict:}\quad & \widehat{\ve{S}}_t = \ve{D}_t \ve{S}_{t-1},
  & \widehat{\ve{P}}_t &= \ve{D}_t \ve{P}_{t-1} \ve{D}_t^{\tr} + \ve{Q}_t,\\[2pt]
  \text{update:}\quad  & \ve{S}_t = \widehat{\ve{S}}_t
  + \ve{\kappa}_t\,(\ve{v}_t - \widehat{\ve{S}}_t^{\tr}\ve{k}_t)^{\tr},
  & \ve{P}_t &= (\ve{I} - \ve{\kappa}_t \ve{k}_t^{\tr})\widehat{\ve{P}}_t,
\end{aligned}
\label{eq:mem-kf}
\end{equation}
with gain $\ve{\kappa}_t = \widehat{\ve{P}}_t \ve{k}_t/(r_t + \ve{k}_t^{\tr}\widehat{\ve{P}}_t \ve{k}_t)$. In the static-state limit
($\ve{D}_t=\ve{I}$, $\ve{Q}_t=0$) the estimate has the closed weighted-least-squares form
\begin{equation}
  \ve{S}_t = \ve{P}_t \sum_{\tau\le t}\gamma_{t,\tau}\,\frac{1}{r_\tau}\ve{k}_\tau \ve{v}_\tau^{\tr},
  \qquad
  \ve{P}_t = \Big(\sum_{\tau\le t}\gamma_{t,\tau}\,\frac{1}{r_\tau}
  \ve{k}_\tau \ve{k}_\tau^{\tr}+\ve{P}_0^{-1}\Big)^{-1},
  \label{eq:mem-wls}
\end{equation}
which \eqref{eq:mem-kf} computes online; the gain $\ve{\kappa}_t$ is a data- and state-dependent
write strength that replaces the delta rule's fixed $\beta_t \ve{k}_t$.

\paragraph{Why it resists parallelization.}
The delta-rule family trains in parallel because its per-token update is an \emph{affine}
map of the state, $\ve{S}_t=\ve{A}_t \ve{S}_{t-1}+\ve{b}_t$, whose coefficients
$\ve{A}_t=(\ve{I}-\beta_t \ve{k}_t \ve{k}_t^{\tr})\ve{D}_t$ and
$\ve{b}_t=\beta_t \ve{k}_t \ve{v}_t^{\tr}$ depend on the \emph{inputs
alone}; a chunk of such maps composes into dense matmuls (the WY / chunked scan). The
optimal filter \eqref{eq:mem-kf} breaks this on two counts. First, the covariance obeys a
\emph{Riccati recursion}
$\ve{P}_t=(\ve{I}-\ve{\kappa}_t \ve{k}_t^{\tr})(\ve{D}_t \ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t)$ that is
nonlinear (rational) in $\ve{P}_{t-1}$ and has no closed-form chunk composition --- it must be
iterated token by token. Second, the gain $\ve{\kappa}_t$ depends on $\ve{P}_{t-1}$, so the state
coefficient $\ve{A}_t$ depends on the \emph{accumulated} covariance rather than the current
input, and the recurrence is no longer a fixed linear scan. (Letting $\ve{D}_t$ or $\ve{Q}_t$ depend
on $\ve{S}_{t-1}$ compounds this, making even the predict step nonlinear.) On top of the
sequential dependence, the covariance is $d_k\times d_k$: $O(d_k^2)$ state and work per
step, versus the delta family's $O(d_k d_v)$ matmuls.

\paragraph{Parallelizable relaxations.}
Parallel training needs the per-token coefficients to be either input-only or computable by
an \emph{associative scan} (a prefix operation). Several general relaxations recover this
while keeping some of the second-order/state dependence the delta family discards:
\begin{itemize}[leftmargin=1.4em,itemsep=3pt,topsep=2pt]
  \item \textbf{Structured covariance.} Restrict $\ve{P}_t$ to a cheap form. A diagonal
  $\ve{P}_t=\diag(\ve{p}_t)$ decouples the Riccati into $d_k$ scalar recursions ($O(d_k)$ rather than
  $O(d_k^2)$); a scalar $\ve{P}_t=b_t \ve{I}$ collapses it to one recursion and a per-step scalar gain
  $\beta_t=b_t/(r_t+b_t)$ for normalized keys; a diagonal-plus-low-rank $\ve{P}_t$ interpolates.
  \item \textbf{Gain from prefix statistics.} Replace the exact Riccati by a gain that is a
  pointwise function of \emph{cumulative} quantities a scan can produce --- e.g.\ a running
  key energy $\sum_{\tau\le t}\gamma_{t,\tau}\ve{k}_\tau \ve{k}_\tau^{\tr}$ (or its diagonal), or an
  EMA of the innovation. The prefix-sum is parallel; the nonlinearity is applied pointwise.
  \item \textbf{Chunk-wise freezing.} Hold the covariance/gain (and any state-dependent
  transition) fixed within a chunk at its chunk-start value, updating it only across chunks.
  Intra-chunk the update is the fixed-gain delta rule (parallel matmuls); inter-chunk a short
  sequential pass carries $\ve{P}$. This interpolates between the delta family (never update) and
  the exact filter (chunk size one), mirroring mini-batch test-time training.
  \item \textbf{Learned surrogate.} Replace the Riccati solution by a small learned map from
  cheap, parallelizable features (the input, prefix statistics, the innovation) to the gain
  and transition --- a parametric stand-in for the second-order recursion, trained end to
  end. State-dependent gates are one member of this class.
\end{itemize}
Each trades fidelity to the Kalman optimum for parallelism and cost; the delta family is the
extreme that freezes the covariance entirely, and richer choices move back toward the optimum
along one of these axes.


% =============================================================================
\subsection{Associative memory as recursive least squares}
\label{app:mem-rls}

\subsubsection{Static memory: recursive least squares}
\label{app:mem-rls-static}

\paragraph{Online least-squares objective.}
Dropping the drift ($\ve{D}_t=\ve{I}$, $\ve{Q}_t=0$) reduces the filter of \S\ref{app:mem-kf} to estimating
a \emph{fixed} memory from streaming observations --- ridge-regularized, recency-weighted
least squares,
\begin{equation}
  \ve{S}_t = \argmin_{\ve{S}}\ \tfrac12\sum_{\tau\le t}\lambda^{t-\tau}
  \big\|\ve{S}^{\tr}\ve{k}_\tau-\ve{v}_\tau\big\|_2^2
        + \tfrac12\lambda^{t}\mu\|\ve{S}\|_F^2 ,
  \label{eq:rls-obj}
\end{equation}
with forgetting factor $\lambda\in(0,1]$ and ridge $\mu>0$ (a decaying prior); this is the
static-state case of \eqref{eq:mem-wls}.

\paragraph{The RLS recursion.}
Zeroing the gradient gives the normal equations $\ve{C}_t \ve{S}_t = \ve{B}_t$ with the \emph{information
matrix} and target
\begin{equation}
  \ve{C}_t = \sum_{\tau\le t}\lambda^{t-\tau}\ve{k}_\tau \ve{k}_\tau^{\tr}
  + \lambda^{t}\mu \ve{I},
  \qquad
  \ve{B}_t = \sum_{\tau\le t}\lambda^{t-\tau}\ve{k}_\tau \ve{v}_\tau^{\tr},
  \label{eq:rls-info}
\end{equation}
both accumulating linearly, $\ve{C}_t=\lambda \ve{C}_{t-1}+\ve{k}_t \ve{k}_t^{\tr}$ and
$\ve{B}_t=\lambda \ve{B}_{t-1}+\ve{k}_t \ve{v}_t^{\tr}$. With $\ve{P}_t=\ve{C}_t^{-1}$, the Sherman--Morrison identity gives
the classical \emph{recursive least squares} update
\begin{equation}
\begin{aligned}
  \ve{g}_t &= \frac{\ve{P}_{t-1}\ve{k}_t}
  {\lambda+\ve{k}_t^{\tr}\ve{P}_{t-1}\ve{k}_t}, \\[3pt]
  \ve{S}_t &= \ve{S}_{t-1}+\ve{g}_t\,
  (\ve{v}_t-\ve{S}_{t-1}^{\tr}\ve{k}_t)^{\tr}, \\[3pt]
  \ve{P}_t &= \lambda^{-1}\big(\ve{P}_{t-1}-\ve{g}_t \ve{k}_t^{\tr}\ve{P}_{t-1}\big),
\end{aligned}
\label{eq:rls-rec}
\end{equation}
the exact minimizer of \eqref{eq:rls-obj} computed online.

\paragraph{RLS is online Newton.}
Since $\ve{g}_t=\ve{P}_t \ve{k}_t$, the update \eqref{eq:rls-rec} is a \emph{preconditioned} gradient step on
the per-token recall loss $\mc L_t(\ve{S})=\tfrac12\|\ve{S}^{\tr}\ve{k}_t-\ve{v}_t\|_2^2$,
\begin{equation}
  \ve{S}_t = \ve{S}_{t-1} - \ve{P}_t\,\nabla_{\!\ve{S}}\mc L_t(\ve{S}_{t-1}),
  \qquad \ve{P}_t = \ve{C}_t^{-1},
  \label{eq:rls-newton}
\end{equation}
whose preconditioner $\ve{P}_t$ is the inverse accumulated Hessian --- so RLS is \emph{online
Newton} (a natural-gradient step). The delta-rule family replaces $\ve{P}_t$ by a fixed
$\beta_t \ve{I}$: \emph{online SGD}, the first-order ($\ve{P}=\ve{I}$) approximation of this second-order
($\ve{P}=\ve{C}^{-1}$) optimum, with GDN/KDA adding an $\ell_2$ shrinkage (the decay) on top.

\paragraph{Information form: a linear accumulator.}
The obstruction to parallel training is \emph{not} the accumulation: by \eqref{eq:rls-info}
the information matrix $\ve{C}_t$ (and $\ve{B}_t$) is a decayed cumulative sum --- a prefix scan, hence
parallelizable. The nonlinearity is confined to the per-step inverse $\ve{P}_t=\ve{C}_t^{-1}$ (or
$\ve{S}_t=\ve{C}_t^{-1}\ve{B}_t$). Carrying $\ve{C}_t$ rather than $\ve{P}_t$ --- the \emph{information} form --- is
therefore the parallel-friendly representation: the state accumulates linearly, and only the
read-out inverts.

\subsubsection{Drifting memory: dynamic least squares}
\label{app:mem-rls-drift}

Adding the drift $\ve{D}_t$ changes the optimization problem qualitatively: the unknown is no
longer one fixed memory $\ve{S}$, but a trajectory of latent memories
$\ve{S}_{0:t}=(\ve{S}_0,\ldots,\ve{S}_t)$ coupled by the process model. The batch MAP problem
corresponding to \eqref{eq:mem-ssm} is
\begin{equation}
  \{\ve{S}_\tau^\star\}_{\tau=0}^t
  =
  \argmin_{\ve{S}_0,\ldots,\ve{S}_t}
  \frac12\|\ve{S}_0\|_{\ve{P}_0^{-1}}^2
  +\frac12\sum_{\tau=1}^t
  \|\ve{S}_\tau-\ve{D}_\tau\ve{S}_{\tau-1}\|_{\ve{Q}_\tau^{-1}}^2
  +\frac12\sum_{\tau=1}^t
  \frac{1}{r_\tau}
  \|\ve{S}_\tau^{\tr}\ve{k}_\tau-\ve{v}_\tau\|_2^2 ,
  \label{eq:drift-map-obj}
\end{equation}
where $\|\ve{A}\|_{\ve{M}}^2:=\Trace(\ve{A}^{\tr}\ve{M}\ve{A})$. The first term is the prior,
the second penalizes deviations from the deterministic drift
$\ve{S}_\tau\approx\ve{D}_\tau\ve{S}_{\tau-1}$, and the third is the observation loss at the
current key. Thus $\ve{Q}_\tau$ controls how much the memory is allowed to deviate from the
drifted previous memory: small $\ve{Q}_\tau$ enforces nearly deterministic transport, while
large $\ve{Q}_\tau$ allows abrupt changes.

The online filter is the one-step version of this dynamic least-squares problem. Given the
posterior at $t-1$, first predict
\begin{equation}
  \widehat{\ve{S}}_t=\ve{D}_t\ve{S}_{t-1},
  \qquad
  \widehat{\ve{P}}_t=\ve{D}_t\ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t .
  \label{eq:drift-filter-predict}
\end{equation}
Then the filtered memory is the minimizer
\begin{equation}
  \ve{S}_t
  =
  \argmin_{\ve{S}}
  \frac12\|\ve{S}-\widehat{\ve{S}}_t\|_{\widehat{\ve{P}}_t^{-1}}^2
  +\frac{1}{2r_t}
  \|\ve{S}^{\tr}\ve{k}_t-\ve{v}_t\|_2^2 .
  \label{eq:drift-filter-obj}
\end{equation}
Solving \eqref{eq:drift-filter-obj} gives the Kalman residual write in
\eqref{eq:mem-kf}. Compared with static RLS, the normal equations no longer accumulate around a
single $\ve{S}$; the batch objective is block-tridiagonal in the trajectory
$\ve{S}_{0:t}$, and the online solution carries the predicted covariance
$\widehat{\ve{P}}_t$ forward through the drift.

% =============================================================================
\subsection{Isotropic Kalman Linear Attention}
\label{app:iso-kla}

Consider the most restrictive covariance family, the isotropic family
\begin{equation}
  \ve{P}_t\in\{b\ve{I}: b\in\R_+\},
  \qquad
  \ve{P}_t=b_t\ve{I},
  \qquad
  \ve{C}_t=\ve{P}_t^{-1}=c_t\ve{I},
  \quad c_t=b_t^{-1},
  \label{eq:iso-cov-family}
\end{equation}
which keeps a single scalar uncertainty $b_t$ (equivalently a scalar information $c_t=b_t^{-1}$)
for the entire key space. We derive the resulting update by specializing the diagonal KLA
recursion to this family with an independently parameterized additive process noise: predict
the scalar uncertainty through the transition, apply the scalar measurement update, and read
off the induced gain. The price of using an independent $q_t$ is that the information recursion
is no longer affine in $c_t$; it is a M\"obius map, which is still associative through a
$2\times2$ matrix scan.

\paragraph{Prediction.}
With a diagonal transition $\ve{D}_t=\diag(\ve{\alpha}_t)$ and additive process covariance
$q_t\ve{I}$, the covariance prediction is
$\widehat{\ve{P}}_t=b_{t-1}\ve{D}_t^2+q_t\ve{I}$. Projecting this diagonal matrix back to the
isotropic family by matching average variance gives
\begin{equation}
  \widehat{b}_t
  =
  a_t b_{t-1}+q_t,
  \qquad
  a_t=\frac{1}{d_k}\sum_{i=1}^{d_k}\ve{\alpha}_{t,i}^2,
  \qquad
  \widehat{c}_t=\frac{c_{t-1}}{a_t+q_t c_{t-1}}.
  \label{eq:iso-kla-predict}
\end{equation}
If $\ve{\alpha}_t$ is uniform across channels, this projection is exact. If it is channel-wise,
the memory transition can still be KDA's $\ve{D}_t=\diag(\ve{\alpha}_t)$, but the uncertainty
used by the scalar gain sees only the average contraction $a_t$. A diagonal process covariance
could be used instead by replacing $q_t$ with its average diagonal value.

\paragraph{Information update and associative scan.}
In the diagonal information form the Kalman measurement update is additive: the posterior
information is the predicted information plus the diagonal of the observation's Fisher
information $\ve{k}_t\ve{k}_t^{\tr}/r_t$, i.e.\ componentwise
\begin{equation}
  \ve{c}_{t,i}
  =
  \widehat{\ve{c}}_{t,i}+\frac{\ve{k}_{t,i}^2}{r_t},
  \qquad i=1,\dots,d_k .
  \label{eq:iso-kla-diag-info-update}
\end{equation}
We collapse this posterior information vector to the isotropic family using the trace, i.e.\
the total Fisher information in the observed key direction. The predicted covariance is already
isotropic, so its scalar information is $\widehat{c}_t$; adding the trace of the measurement
information and using $\sum_i\ve{k}_{t,i}^2=\|\ve{k}_t\|_2^2$ gives
\begin{equation}
  c_t
  =
  \widehat{c}_t+\sum_{i=1}^{d_k}\frac{\ve{k}_{t,i}^2}{r_t}
  =
  \widehat{c}_t+\frac{\|\ve{k}_t\|_2^2}{r_t},
  \qquad
  c_0=\mu,
  \label{eq:iso-kla-info-update}
\end{equation}
which is the scalar measurement update. This trace-scaled isotropic projection avoids the extra
per-coordinate factor $1/d_k$: for normalized keys, each token contributes information $1/r_t$
to the scalar uncertainty tracker.

Combining the prediction and update gives a linear-fractional recurrence in the previous
information state. With
\begin{equation}
  u_t=\frac{\|\ve{k}_t\|_2^2}{r_t},
  \label{eq:iso-kla-u}
\end{equation}
we have
\begin{equation}
  c_t
  =
  \frac{c_{t-1}}{a_t+q_t c_{t-1}}+u_t
  =
  \frac{(1+u_t q_t)c_{t-1}+u_t a_t}{q_t c_{t-1}+a_t}.
  \label{eq:iso-kla-mobius}
\end{equation}
Thus each token defines a M\"obius map
\begin{equation}
  c_t=f_t(c_{t-1}),
  \qquad
  f_t(c)=\frac{A_t c+B_t}{C_t c+D_t},
  \qquad
  \ve{M}_t=
  \begin{bmatrix}
    A_t & B_t\\
    C_t & D_t
  \end{bmatrix}
  =
  \begin{bmatrix}
    1+u_t q_t & u_t a_t\\
    q_t & a_t
  \end{bmatrix}.
  \label{eq:iso-kla-mobius-matrix}
\end{equation}
M\"obius maps compose by matrix multiplication:
\begin{equation}
  f_t\circ f_{t-1}\circ\cdots\circ f_1
  \quad\Longleftrightarrow\quad
  \ve{M}_t\ve{M}_{t-1}\cdots\ve{M}_1 .
  \label{eq:iso-kla-mobius-compose}
\end{equation}
Equivalently, in homogeneous coordinates $c_t=n_t/d_t$ and
$[n_t,d_t]^{\tr}=\ve{M}_t[n_{t-1},d_{t-1}]^{\tr}$. Therefore all prefix information states can
be computed by an associative scan over $2\times2$ matrices, with $c_0=\mu$.

\paragraph{Gain.}
The predicted uncertainty is now isotropic, $\widehat{\ve{P}}_t=\widehat{b}_t\ve{I}$ with
$\widehat{b}_t=1/\widehat{c}_t$ from \eqref{eq:iso-kla-predict}. Substituting into the Kalman gain
$\ve{\kappa}_t=\widehat{\ve{P}}_t\ve{k}_t/(r_t+\ve{k}_t^{\tr}\widehat{\ve{P}}_t\ve{k}_t)$, the
covariance acts as a scalar and the gain collapses to the fixed delta-rule direction:
\begin{equation}
  \ve{\kappa}_t
  =
  \frac{\widehat{b}_t}{r_t+\widehat{b}_t\|\ve{k}_t\|_2^2}\,\ve{k}_t
  \equiv
  \beta_t\ve{k}_t .
  \label{eq:iso-kla-gain}
\end{equation}

\paragraph{Memory update.}
With $\beta_t$ from \eqref{eq:iso-kla-gain}, the memory update is exactly the
fixed-gain residual form
\begin{equation}
  \ve{S}_t
  =
  (\ve{I}-\beta_t\ve{k}_t\ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}
  +\beta_t\ve{k}_t\ve{v}_t^{\tr}.
  \label{eq:iso-kla-memory}
\end{equation}
Thus isotropic KLA can adapt the scalar write strength $\beta_t$ through the Kalman uncertainty
scan, but --- keeping only one scalar for the entire key space --- it cannot assign different
write strengths to different key channels; an anisotropic gain is recovered only by relaxing to
the diagonal family.

\refstepcounter{definition}\label{prop:isotropic-kalman-linear-attention}
\begin{tcolorbox}[
  title={Proposition~\thedefinition: Isotropic Kalman Linear Attention},
  colback=blue!7,
  colbacktitle=blue!22,
  colframe=blue!45!black,
  coltitle=black,
  fonttitle=\bfseries,
  boxrule=0.4pt,
  arc=1pt,
  left=5pt,
  right=5pt,
  top=5pt,
  bottom=5pt
]
Isotropic Kalman Linear Attention is the scalar-uncertainty specialization of KLA. Given
token representation $\ve{x}_t$, key--value pair $(\ve{k}_t,\ve{v}_t)$, and previous memory
$\ve{S}_{t-1}$, the layer parameterizes a diagonal memory transition, an additive process noise,
and a token-dependent observation noise,
\begin{equation}
\begin{aligned}
  \ve{\alpha}_t &= \sigma(\ve{W}_{\alpha}\ve{x}_t+\ve{b}_{\alpha}),
  &
  q_t &= q_{\min}+\operatorname{softplus}(\ve{w}_{q}^{\tr}\ve{x}_t+b_q),
  &
  r_t &= r_{\min}+\operatorname{softplus}(\ve{w}_{r}^{\tr}\ve{x}_t+b_r),
  \\[2pt]
  a_t &= \frac{1}{d_k}\sum_i\ve{\alpha}_{t,i}^2,
  &
  \ve{D}_t &= \diag(\ve{\alpha}_t).
\end{aligned}
\label{eq:iso-kla-box-parameterization}
\end{equation}
It then computes scalar prefix information statistics by a M\"obius scan:
\begin{equation}
  u_t=\frac{\|\ve{k}_t\|_2^2}{r_t},
  \qquad
  \ve{M}_t=
  \begin{bmatrix}
    1+u_t q_t & u_t a_t\\
    q_t & a_t
  \end{bmatrix},
  \qquad
  \begin{bmatrix}n_t\\ d_t\end{bmatrix}
  =
  \ve{M}_t
  \begin{bmatrix}n_{t-1}\\ d_{t-1}\end{bmatrix},
  \qquad
  c_t=\frac{n_t}{d_t},
  \qquad
  \begin{bmatrix}n_0\\ d_0\end{bmatrix}
  =
  \begin{bmatrix}\mu\\ 1\end{bmatrix}.
  \label{eq:iso-kla-box-mobius}
\end{equation}
The gain at token $t$ uses the exclusive prefix state $c_{t-1}$:
\begin{equation}
\begin{aligned}
  \widehat{c}_t
  &= \frac{c_{t-1}}{a_t+q_t c_{t-1}},
  &
  \widehat{b}_t
  &= \frac{1}{\widehat{c}_t},
  &
  \beta_t
  &= \frac{\widehat{b}_t}{r_t+\widehat{b}_t\|\ve{k}_t\|_2^2}.
\end{aligned}
\label{eq:iso-kla-box-gain}
\end{equation}
With $\beta_t$ available before the memory scan, the memory update is the fixed-direction
affine scan
\begin{equation}
  \ve{S}_t
  =
  (\ve{I}-\beta_t\ve{k}_t\ve{k}_t^{\tr})\ve{D}_t\ve{S}_{t-1}
  +\beta_t\ve{k}_t\ve{v}_t^{\tr}.
  \label{eq:iso-kla-box-memory}
\end{equation}
Thus ISO-KLA preserves the Kalman idea of adapting write strength from uncertainty, but the
isotropic covariance restriction forces the gain direction to be $\ve{k}_t$.
\end{tcolorbox}

Therefore ISO-KLA sits between fixed-gain delta rules and diagonal KLA: it keeps a
Kalman-style scalar uncertainty and token-dependent observation noise, but because
$\ve{P}_t$ is isotropic, it always writes along $\ve{k}_t$ with one scalar strength. Diagonal
KLA strictly generalizes this by replacing the scalar $b_t$ with a vector
$\ve{p}_t$, allowing the gain to be anisotropic.

% =============================================================================
\subsection{Diagonal information: from the covariance recursion to the M\"obius scan}
\label{app:diag-info}

This section derives, step by step, the predicted uncertainty $\widehat{\ve{p}}_t$ of
\S\ref{sec:kla} from the diagonal covariance recursion. With an independently parameterized
process noise, the information update is a M\"obius map; this remains parallelizable because
M\"obius maps compose by $2\times2$ matrix multiplication. All products, inverses, and ratios
are coordinatewise; we write a single channel and drop the index.

\paragraph{Diagonal state.}
Restrict the key-space covariance of the associative-memory filter \eqref{eq:mem-kf} to
diagonal, $\ve{P}_t=\diag(\ve{p}_t)$, with transition $\ve{D}_t=\diag(\ve{\alpha}_t)$ and process
noise $\ve{Q}_t=\diag(\ve{q}_t)$. Per channel, track the variance $p_t$ and its reciprocal, the
information $c_t=1/p_t$.

\paragraph{Step 1 --- predict (the master recursion).}
The covariance predict $\widehat{\ve{P}}_t=\ve{D}_t\ve{P}_{t-1}\ve{D}_t^{\tr}+\ve{Q}_t$ preserves
diagonality, and per channel is
\begin{equation}
  \boxed{\;\widehat{p}_t=\alpha_t^2\,p_{t-1}+q_t\;}
  \label{eq:app-predict}
\end{equation}
(the diagonal form of \eqref{eq:diag-covariance-prediction}). This is the \emph{only} exact
predicted variance; every expression below is \eqref{eq:app-predict} written in different
variables.

\paragraph{Step 2 --- update, in information form.}
The measurement step of the information filter adds the observation's information,
$\ve{C}_t=\widehat{\ve{C}}_t+\ve{k}_t\ve{k}_t^{\tr}/r_t$; keeping the diagonal (the diagonal
restriction) gives, per channel,
\begin{equation}
  c_t=\widehat{c}_t+\frac{k_t^2}{r_t},
  \qquad
  \widehat{c}_t:=\frac{1}{\widehat{p}_t}.
  \label{eq:app-update}
\end{equation}
So $\widehat{c}_t$ is \emph{defined} as $1/\widehat{p}_t$; it is not a separate quantity.

\paragraph{Step 3 --- the recursion.}
Insert \eqref{eq:app-predict} into $\widehat{c}_t=1/\widehat{p}_t$ and use $p_{t-1}=1/c_{t-1}$:
\begin{equation}
  \widehat{c}_t
  =\frac{1}{\alpha_t^2\,p_{t-1}+q_t}
  =\frac{1}{\alpha_t^2/c_{t-1}+q_t}
  =\frac{c_{t-1}}{\alpha_t^2+q_t\,c_{t-1}},
  \label{eq:app-chat}
\end{equation}
so the full posterior recursion is
\begin{equation}
  c_t=\frac{c_{t-1}}{\alpha_t^2+q_t\,c_{t-1}}+\frac{k_t^2}{r_t}.
  \label{eq:app-cfull}
\end{equation}
For a general $q_t$ this is a \emph{M\"obius} (linear-fractional) map of $c_{t-1}$,
$c_t=(A_tc_{t-1}+B_t)/(C_tc_{t-1}+D_t)$, with
\begin{equation}
  \begin{bmatrix}A_t&B_t\\[2pt] C_t&D_t\end{bmatrix}
  =\begin{bmatrix}1+\tfrac{k_t^2}{r_t}\,q_t & \tfrac{k_t^2}{r_t}\,\alpha_t^2\\[3pt] q_t & \alpha_t^2\end{bmatrix}.
  \label{eq:app-mobius}
\end{equation}
M\"obius maps compose by matrix multiplication, so $c_T$ is the action of the product
$\ve{M}_T\cdots\ve{M}_1$ on $c_0$. When $q_t$ is input-only the matrices are input-only, the
product is associative, and it parallelizes as the scan in \eqref{eq:diag-info-scan} (at higher
constant cost than an affine scan, and needing per-chunk renormalization since the entries can
span a wide range). This is the diagonal-information recursion used by KLA with freely
parameterized additive process noise.

\paragraph{Step 4 --- the special choice that linearizes the scan.}
Equation~\eqref{eq:app-cfull} is affine in $c_{t-1}$ iff the product $q_t\,c_{t-1}$ is
state-independent, i.e.\ iff $q_t\propto p_{t-1}$. Writing this as
\begin{equation}
  q_t=(\lambda_t^{-1}-\alpha_t^2)\,p_{t-1}
  \qquad\Longleftrightarrow\qquad
  \lambda_t=\frac{1}{\alpha_t^2+q_t/p_{t-1}}
  \label{eq:app-lambda}
\end{equation}
(the identity \eqref{eq:app-lambda}) makes $\alpha_t^2+q_t\,c_{t-1}=\lambda_t^{-1}$, so
\eqref{eq:app-chat}--\eqref{eq:app-cfull} collapse to the affine scan
\begin{equation}
  \widehat{c}_t=\lambda_t\,c_{t-1},
  \qquad
  c_t=\lambda_t\,c_{t-1}+\frac{k_t^2}{r_t},
  \label{eq:app-affine}
\end{equation}
which is the cheaper affine-scan restriction. Substituting the \emph{same} $q_t$ back into the master
recursion \eqref{eq:app-predict},
\begin{equation}
  \widehat{p}_t=\alpha_t^2\,p_{t-1}+q_t=\lambda_t^{-1}\,p_{t-1}=\frac{1}{\widehat{c}_t},
  \label{eq:app-phat}
\end{equation}
i.e.\ \eqref{eq:prefix-diag-uncertainty}. The transition term $\alpha_t^2$ is not dropped; it is
exactly cancelled by this $q_t$. The small prior $\mu$ enters only through the initialization
$c_0=\mu$, which keeps the information state positive at the start of the scan.

\paragraph{Recovering KDA.}
KLA reduces to KDA when the transition remains KDA's diagonal transition and the predicted
uncertainty becomes isotropic. In the diagonal information form, this means
\begin{equation}
  \ve{D}_t=\diag(\ve{\alpha}_t),
  \qquad
  \widehat{\ve{p}}_t=b_t\ve{1}
  \quad\Longleftrightarrow\quad
  \widehat{\ve{c}}_t=b_t^{-1}\ve{1}
  \label{eq:app-kda-isotropic-condition}
\end{equation}
for some scalar $b_t>0$. Since KLA parameterizes an additive process noise and uses
$\widehat{p}_{t,i}=\alpha_{t,i}^2p_{t-1,i}+q_{t,i}$, the isotropic condition is
\begin{equation}
  \alpha_{t,i}^2p_{t-1,i}+q_{t,i}
  =
  b_t,
  \qquad\text{equivalently}\qquad
  q_{t,i}=b_t-\alpha_{t,i}^2p_{t-1,i},
  \label{eq:app-kda-q-condition}
\end{equation}
for every channel $i$, with the validity condition $q_{t,i}\ge 0$. Under
\eqref{eq:app-kda-isotropic-condition}, the KLA gain collapses to
\begin{equation}
  \ve{\kappa}_t
  =
  \frac{b_t}{r_t+b_t\|\ve{k}_t\|_2^2}\,\ve{k}_t
  \equiv \beta_t\ve{k}_t,
  \label{eq:app-kda-gain-collapse}
\end{equation}
and the memory update becomes the vanilla KDA residual update
\begin{equation}
  \ve{S}_t
  =
  (\ve{I}-\beta_t\ve{k}_t\ve{k}_t^{\tr})\diag(\ve{\alpha}_t)\ve{S}_{t-1}
  +\beta_t\ve{k}_t\ve{v}_t^{\tr}.
  \label{eq:app-kda-recovered}
\end{equation}
For normalized keys, $\beta_t=b_t/(r_t+b_t)$.

\paragraph{Consequences.}
\begin{itemize}[leftmargin=1.4em,itemsep=3pt,topsep=2pt]
  \item Equation~\eqref{eq:app-predict}, $\widehat{p}_t=\alpha_t^2 p_{t-1}+q_t$, is the master
  recursion; \eqref{eq:app-chat} is the same recursion written in information variables.
  \item $\widehat{p}_t=1/\widehat{c}_t$ is not an additional modeling term. Since
  $\widehat{c}_t=c_{t-1}/(\alpha_t^2+q_t c_{t-1})$, its inverse is exactly
  $\alpha_t^2p_{t-1}+q_t$. Adding $q_t$ again would double-count the process noise.
  \item Free, input-only, $p$-independent $q_t$ gives the M\"obius scan
  \eqref{eq:app-mobius}; the older $\lambda$-form $q_t\propto p_{t-1}$ is the special case that
  collapses this to the affine scan \eqref{eq:app-affine}.
\end{itemize}




\end{document}
