\RequirePackage[svgnames,table]{xcolor}

\documentclass[11pt,letterpaper,logo]{yalearxiv}

\input{preamble}
\input{command}


\newtcolorbox{simpleElegantQuote}{
    colback=AliceBlue!50!White,
    colframe=RoyalBlue!75!Black,
    boxrule=0.5pt,
    arc=2mm,
    boxsep=4pt,
    left=10pt, right=10pt,
    top=8pt, bottom=8pt,
    fontupper=\itshape,
}

\title{HyperDeltaNet: Manifold-Constrained Memory for Linear Attention}


\runningtitle{HyperDeltaNet}

\usepackage{fontawesome5}
\definecolor{myorange}{RGB}{250,99,57}
\newcommand{\orgfire}{\textcolor{myorange}{\faFire*}}
\newcommand{\trimkv}{TrimKV\textsuperscript{ \orgfire}}
\newcommand{\dbtrimkv}{DBTrimKV\textsuperscript{ \orgfire}}

\definecolor{yaleblue}{RGB}{0,58,112}
\definecolor{jpmcbrown}{RGB}{153,108,72}
\newcommand{\yale}{%
  \textsuperscript{%
    {\usefont{T1}{pbk}{m}{n}\textcolor{yaleblue}{\textbf{Y}}}%
  }%
}

\author{Ngoc Bui\yale, Rex Ying\yale\\
\yale Department of Computer Science, Yale University\\
\faGithub~\textbf{Source Code:} \href{https://github.com/ngocbh/hyper-delta-net}{\texttt{https://github.com/ngocbh/hyper-delta-net}}
}

\hypersetup{colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black,
            urlcolor=blue!50!black}

\newcommand{\alphaa}{\alpha}
\newcommand{\betab}{\beta}
\newcommand{\neff}[1][]{n_{\text{eff}#1}}

% Provisional / preliminary number tag
\newcommand{\prelim}[1]{\textcolor{red!70!black}{#1$^{\dagger}$}}

% --- draft-local notation ---
\newcommand{\smax}{\sigma_{\max}}
\newcommand{\Tcarry}{T}          % horizontal carry operator T_t
\newcommand{\Hres}{H_{\mathrm{res}}}
\newcommand{\Hpre}{H_{\mathrm{pre}}}
\newcommand{\Hpost}{H_{\mathrm{post}}}


\begin{document}


\begin{abstract}
\vspace{-1mm}
{\centering\section*{Abstract}}
Linear-attention models compress an unbounded KV cache into a fixed-size recurrent
state, so the central design question is how to \emph{edit} that compressed memory
without scrambling stored associations. Gated DeltaNet-2 (GDN-2) advances this by
\emph{decoupling} the erase and write operations into separate channel-wise gates,
setting a new frontier for recurrent long-context retrieval. We show that this
decoupling carries a hidden, and in fact \emph{unavoidable}, cost: the per-step
state-transition operator becomes \emph{non-normal}, so its largest singular value
strictly exceeds one whenever the erase gate is genuinely channel-wise. We prove a
tight closed form for this spectral norm and show equality with one holds
\emph{only} when the model collapses back to a scalar (KDA-style) erase. GDN-2 thus
buys expressiveness by selling step-wise non-expansiveness --- precisely in the
long-memory regime where it claims its gains. We recast the memory recurrence as the
\emph{time-axis} analog of the residual carry in hyper-connections, which lets us
import the manifold-constraint principle of mHC. The impossibility result forces
exactly two remedies --- \emph{damp} the transition onto the non-expansive set, or
\emph{enrich} it with orthogonal structure --- which we develop as HyperDeltaNet.
We further generalize the state into multiple doubly-stochastically mixed streams,
and prove that constrained per-stream transitions compose with mass-preserving
mixing into a memory operator that is non-expansive by construction. We lay out an
experimental program to test whether the predicted instability is empirically
consequential and whether the fix improves long-context retrieval.
\end{abstract}


\maketitle


% =============================================================================
\section{Introduction}
\label{sec:intro}

Linear attention replaces the growing key--value cache of softmax attention with a
fixed-size recurrent state $S_t \in \R^{d_k \times d_v}$, trading unbounded memory
for constant-time, constant-memory inference \citep{katharopoulos2020transformers}.
The price is that every token must be \emph{written into} a compressed store that
cannot grow: the model must continually decide what to overwrite, what to preserve,
and what to forget. The delta-rule family \citep{schlag2021linear,yang2024parallelizing}
made this an explicit, associative edit --- remove the old value bound to a key
before writing the new one \citep{schlag2021linear,yang2024parallelizing} --- and
successive refinements
(Gated DeltaNet \citep{yang2024gdn}, Kimi Delta Attention) added gating and
channel-wise decay for finer control.

Gated DeltaNet-2 (GDN-2) \citep{hatamizadeh2026gdn2} pushes this line furthest by
\emph{decoupling} two decisions that all prior delta-rule models had tied to a
single scalar gate $\beta_t$: a channel-wise \emph{erase} gate $b_t$ that selects
which key-side coordinates of the state to remove, and a channel-wise \emph{write}
gate $w_t$ that selects which value-side coordinates to commit. This decoupling is
GDN-2's headline contribution and delivers its strongest results on
interference-heavy multi-key retrieval, exactly the setting that stresses memory
editing.

\paragraph{This paper.} We identify a structural cost of that decoupling and turn it
into a design principle. Our starting observation is that the memory recurrence
$S_t = \Tcarry_t S_{t-1} + \text{write}_t$ is \emph{structurally identical} to the
residual update $x_{l+1} = \Hres\, x_l + F(\cdot)$ of hyper-connections
\citep{zhu2024hyperconnections} and its manifold-constrained variant mHC
\citep{deepseek2025mhc}: the horizontal (cross-\emph{time}) carry $\Tcarry_t$ is the
analog of the depth (cross-\emph{layer}) carry $\Hres$. mHC's central idea is to
constrain $\Hres$ to a norm-controlled manifold so the residual highway neither
explodes nor vanishes across depth. We ask whether the same discipline stabilizes
the memory highway across time --- and find that GDN-2's decoupling is precisely
what makes it necessary.

\paragraph{Contributions.}
\begin{itemize}[leftmargin=1.5em,itemsep=1pt,topsep=2pt]
  \item \textbf{An impossibility result (\S\ref{sec:gap}).} We give a closed form
  for the largest singular value of GDN-2's rank-one erase factor and prove
  $\smax \ge 1$ \emph{always}, with equality iff the erase gate is a per-head
  scalar. Decoupled ($e \not\parallel k$), rank-one, and step-wise non-expansive
  cannot hold simultaneously.
  \item \textbf{Direction A --- constraining the single carry (\S\ref{sec:dirA}).}
  The impossibility forces two escape routes: \emph{damp} the transition back onto
  the non-expansive set (A0, a closed-form drop-in), or \emph{enrich} it with
  orthogonal structure so norm preservation is recovered (A1, Householder/Givens).
  \item \textbf{Direction B --- multi-stream hyper-memory (\S\ref{sec:dirB}).} We
  replace the single state with $n$ streams mixed by a doubly-stochastic matrix at
  chunk boundaries --- hyper-connections along the \emph{memory} axis --- at
  negligible overhead.
  \item \textbf{Composition (\S\ref{sec:compose}).} Non-expansive per-stream carries
  and mass-preserving mixing compose into a memory operator that is provably
  non-expansive, retaining GDN-2's decoupled expressiveness \emph{inside} each
  stream.
  \item \textbf{An honest experimental program (\S\ref{sec:exp}).} We specify the
  measurements that would confirm (or refute) that the predicted non-normal
  amplification is consequential at trained scale, before any kernel work.
\end{itemize}


% =============================================================================
\section{Background}
\label{sec:bg}

\subsection{Two residuals}
\label{sec:two-residuals}

\paragraph{The depth axis: hyper-connections and mHC.}
With $s$ residual streams, a hyper-connection layer updates
\begin{equation}
  x_{l+1} \;=\; \Hres\, x_l \;+\; \Hpost^{\tr}\, F\!\big(\Hpre\, x_l,\, W_l\big),
  \label{eq:hc}
\end{equation}
where $\Hres \in \R^{s\times s}$ mixes streams and $\Hpre,\Hpost$ are separate
read/write maps. mHC \citep{deepseek2025mhc} constrains $\Hres$ to a norm-controlled
manifold --- the Birkhoff polytope (doubly stochastic, via Sinkhorn) or the
orthostochastic set (squared entries of an orthogonal matrix, via Newton--Schulz) ---
so the deep residual highway stays norm-controlled across depth. Crucially, the
\emph{carry} $\Hres$ and the \emph{read/write} maps $\Hpre \neq \Hpost$ are
constrained \emph{separately}.

\paragraph{The time axis: the delta-rule ladder.}
Linear-attention memories share the recurrence $S_t = (\text{carry})\, S_{t-1} +
(\text{write})$, with output $o_t = S_t^{\tr} q_t$. Table~\ref{tab:ladder} places the
main variants on a ladder by their carry operator. Reading down the table, the carry
grows from the identity to a scalar decay, to a symmetric rank-one \emph{projector}
(the delta rule), to gated and channel-wise-decayed versions, and finally to GDN-2's
\emph{non-symmetric} carry.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Model & State update $S_t=\dots$ & Carry on $S_{t-1}$ \\
\midrule
Linear attn & $S_{t-1} + k_t v_t^{\tr}$ & $I$ \\
Mamba-2 & $\alpha_t S_{t-1} + k_t v_t^{\tr}$ & $\alpha_t I$ \\
DeltaNet & $(I - \beta_t k_t k_t^{\tr})S_{t-1} + \beta_t k_t v_t^{\tr}$ & $I - \beta_t k_t k_t^{\tr}$ (\textbf{sym.}) \\
Gated DeltaNet & $\alpha_t(I - \beta_t k_t k_t^{\tr})S_{t-1} + \beta_t k_t v_t^{\tr}$ & $\alpha_t(I - \beta_t k_t k_t^{\tr})$ \\
KDA & $(I - \beta_t k_t k_t^{\tr})D_t S_{t-1} + \beta_t k_t v_t^{\tr}$ & $(I - \beta_t k_t k_t^{\tr})D_t$ \\
\textbf{GDN-2} & $(I - k_t e_t^{\tr})D_t S_{t-1} + k_t z_t^{\tr}$ & $(I - k_t e_t^{\tr})D_t$ (\textbf{non-sym.}) \\
\bottomrule
\end{tabular}
\caption{The delta-rule ladder. $D_t=\diag(\alpha_t)$; for GDN-2,
$e_t = b_t \odot k_t$ (channel-wise erase) and $z_t = w_t \odot v_t$
(channel-wise write). GDN-2 recovers KDA when $b_t = w_t = \beta_t \Onef$, and
Gated DeltaNet when the decay is also scalar.}
\label{tab:ladder}
\end{table}

\subsection{Non-expansive vs.\ expansive carries}
\label{sec:sigmax}

A linear map $M$ is \emph{non-expansive} if $\|Mx\| \le \|x\|$ for all $x$, i.e.\
$\smax(M) = \|M\|_2 \le 1$; it is \emph{expansive} if some direction is stretched
($\smax(M) > 1$). Because the recurrence multiplies the state by $\Tcarry_t$ every
step, old memory passes through the product $\Tcarry_L \cdots \Tcarry_1$; a
non-expansive carry lets old memory only stay or shrink (legal forgetting), while an
expansive carry can let it grow step after step and swamp new writes.

The subtlety is that eigenvalues govern the \emph{asymptotic} fate while singular
values govern a \emph{single step}. A \emph{non-normal} matrix (non-orthogonal
eigenvectors) can have all eigenvalues $\le 1$ yet still stretch transiently:
\begin{equation}
  k=(\tfrac1{\sqrt2},\tfrac1{\sqrt2}),\ \ b=(1,0)
  \ \Rightarrow\ e = b\odot k = (\tfrac1{\sqrt2},0),\ \
  M = I - k e^{\tr} = \begin{bmatrix} 0.5 & 0\\ -0.5 & 1\end{bmatrix},\ \
  \text{eig}=\{0.5,1\},\ \smax\approx 1.14 .
  \label{eq:nonnormal-ex}
\end{equation}
Symmetric matrices are \emph{normal}, so eigenvalues equal singular values --- for
DeltaNet/GDN/KDA, ``eigenvalues $\le 1$'' already implies non-expansive. GDN-2's
decoupling makes the erase non-symmetric and opens the gap.


% =============================================================================
\section{The Non-Expansiveness Gap}
\label{sec:gap}

\subsection{The core mapping}
Define the horizontal carry $\Tcarry_t = (I - k_t e_t^{\tr}) D_t$. Then
$S_t = \Tcarry_t S_{t-1} + k_t z_t^{\tr}$ is structurally identical to the
hyper-connection update~\eqref{eq:hc}. This gives the correspondence that organizes
the paper:
\begin{center}
\emph{$\Tcarry_t$ is the time-axis analog of $\Hres$; the erase gate $b_t$ and write
gate $w_t$ are the analog of $\Hpre$ and $\Hpost$.}
\end{center}
Stability of the long-context recurrence is governed by the spectral behavior of the
cumulative product $\Tcarry_t \cdots \Tcarry_1$ --- exactly what mHC controls for
depth.

\subsection{The impossibility result}

\begin{theorem}[Non-expansiveness boundary]
\label{thm:boundary}
Let $k \in \R^{d_k}$ with $\|k\|=1$, let $b \in \R^{d_k}$, and set $e = b \odot k$,
$M = I - k e^{\tr}$. Write $c = k^{\tr} e = \sum_i b_i k_i^2$ and
$p = \|e\|^2 = \sum_i b_i^2 k_i^2$. Then
\begin{equation}
  \smax(M)^2 = 1 + \lambda_+,
  \qquad
  \lambda_+ = \tfrac12\!\left[(p - 2c) + \sqrt{p\,(p + 4 - 4c)}\,\right] \ge 0 .
  \label{eq:boundary}
\end{equation}
Consequently $\smax(M) \ge 1$ for all $b$, with equality if and only if $p = c^2$,
if and only if $e \parallel k$, if and only if $b$ is a per-head scalar
(the KDA/Gated-DeltaNet case).
\end{theorem}

\begin{proof}[Proof sketch (full proof in Appendix~\ref{app:proof})]
Since $\|k\|^2 = 1$, $M^{\tr}M = I + N$ with $N = -k e^{\tr} - e k^{\tr} + e e^{\tr}$,
whose range lies in $\mathrm{span}\{k,e\}$, so $M^{\tr}M$ has eigenvalue $1$ off this
subspace. Decompose $e = c\,k + f$ with $f \perp k$ and $\|f\|^2 = p - c^2 =: q \ge
0$ (Cauchy--Schwarz with weights $k_i^2$). In the orthonormal basis $\{k, f/\|f\|\}$,
the restriction of $N$ has trace $p - 2c$ and determinant $-q \le 0$, giving
eigenvalues $\lambda_\pm = \tfrac12[(p-2c) \pm \sqrt{(p-2c)^2 + 4q}]$ and
$(p-2c)^2 + 4q = p(p+4-4c)$. The non-positive determinant forces
$\lambda_+ \ge 0 \ge \lambda_-$, so $\smax(M)^2 = 1 + \lambda_+ \ge 1$, with equality
iff $q = 0$, i.e.\ $e \parallel k$.
\end{proof}

\begin{corollary}[Who is safe]
\label{cor:safe}
DeltaNet, Gated DeltaNet, and KDA are step-wise non-expansive: their erase factor
$I - \beta k k^{\tr}$ ($\|k\|=1$, $\beta\in[0,2]$) is symmetric with eigenvalues
$\{1^{(d_k-1)}, 1-\beta\}$, so $\smax = 1$; KDA remains safe under $D_t$ as a product
of two individually non-expansive factors. GDN-2's carry, by
Theorem~\ref{thm:boundary}, has $\smax > 1$ whenever $b$ is genuinely channel-wise.
\end{corollary}

\paragraph{Why this matters where GDN-2 wins.} The decay $D_t$ can mask the extra
gain only when $\alpha$ is small; but the long-memory regime $\alpha \to 1$ --- where
GDN-2 reports its long-context RULER / MK-NIAH gains --- is exactly where masking
vanishes. GDN-2's core innovation (channel-wise $b \Rightarrow e \not\parallel k$)
\emph{necessarily} sacrifices step-wise non-expansiveness, and the model never
constrains it.

\subsection{Eq.~5 as the on-manifold anchor}
The plain delta rule's carry $I - \beta k k^{\tr}$ traces a one-parameter path from
the identity ($\beta=0$) through the orthogonal projector ($\beta=1$) to a
Householder reflection ($\beta=2$) --- sitting inside the norm-controlled orthogonal
family that mHC's orthostochastic/Newton--Schulz variant is built around. It is also
a stable optimizer step: the delta rule is one gradient step on
$\tfrac12\|S^{\tr}k - v\|^2$ (Hessian $k k^{\tr}$), so the carry is
$I - \beta\cdot\text{Hessian}$, stable for $\beta \le 2/\lambda_{\max} = 2$ --- the
same window. GDN-2 stepped off this manifold to gain expressiveness; the rest of the
paper puts each residual carry back on the manifold correct for its axis
(Table~\ref{tab:halves}).

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
 & Constrained carry ($\Hres$-like) & Separated read/write ($\Hpre \neq \Hpost$)\\
\midrule
DeltaNet/KDA & \checkmark\ non-expansive & $\times$\ tied to $k$\\
GDN-2        & $\times$\ $\smax > 1$      & \checkmark\ decoupled\\
\textbf{mHC / HyperDeltaNet} & \checkmark & \checkmark\\
\bottomrule
\end{tabular}
\caption{Each prior model holds half of mHC's discipline. HyperDeltaNet restores
both on the time axis.}
\label{tab:halves}
\end{table}


% =============================================================================
\section{Direction A: Constraining the Single Carry}
\label{sec:dirA}

\paragraph{The dichotomy.} Theorem~\ref{thm:boundary} says a rank-one erase is
non-expansive \emph{iff} $e \parallel k$: one cannot have \emph{decoupled} \textbf{and}
\emph{rank-one} \textbf{and} \emph{non-expansive} at once. Relaxing exactly one of
the latter two yields the two families below. Which manifold is correct depends on
the axis (Table~\ref{tab:manifolds}).

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
Axis & Operator & Correct manifold & Why \\
\midrule
Within a state (time) & transition $\Tcarry_t$ & non-expansive / orthogonal &
forgetting must stay legal (contraction ok, not mass-preserving) \\
Across states (streams) & mixing $H$ & doubly-stochastic / orthostochastic &
forgetting handled per-stream; mixing should preserve mass \\
\bottomrule
\end{tabular}
\caption{The right mHC manifold on each axis. Birkhoff (doubly stochastic) is
\emph{wrong inside} one state (it forbids forgetting) but \emph{right across}
states --- making A and B complementary, not redundant.}
\label{tab:manifolds}
\end{table}

\subsection{A0 --- non-expansive projection (drop-in)}
Per token, compute the two scalars $c = k_t^{\tr} e_t$ and $p = \|e_t\|^2$ ($O(d_k)$),
form $\smax(M_t)$ in closed form via~\eqref{eq:boundary}, and rescale:
\begin{equation}
  \widetilde{\Tcarry}_t = \frac{M_t D_t}{\max\!\big(1,\, \smax(M_t)\big)},
  \qquad M_t = I - k_t e_t^{\tr}.
  \label{eq:a0}
\end{equation}
This guarantees $\|\widetilde{\Tcarry}_t\| \le 1$ every step, kills the non-normal
blow-up, and keeps the erase/write decoupling. The cost is two dot products and a
scalar divide, folded into the decay --- nothing else about GDN-2 changes. It is the
faithful analog of Sinkhorn-projecting $\Hres$, adapted to the rank-one$+$diagonal
structure and to the \emph{non-expansive} set. Variants: \textbf{A0-whole} (divide the
whole transition, adding adaptive decay), \textbf{A0-corr} (damp only the correction
$k e^{\tr}$, softer), \textbf{A0-soft} (a smooth auxiliary penalty
$\lambda\,\mathrm{relu}(\smax-1)^2$ with no forward change), and \textbf{A0-cap-b}
(rescale $b_t$ toward the aligned direction just enough to hit $\smax \le 1+\eps$).

\subsection{A1 --- orthogonal by construction}
Factor the transition into \emph{reorganize} then \emph{forget}:
\begin{equation}
  S_t = D_t\, R_t\, S_{t-1} + k_t z_t^{\tr}, \qquad R_t \text{ orthogonal}.
  \label{eq:a1}
\end{equation}
The single Householder $R_t = I - 2 k_t k_t^{\tr}$ is exactly the delta rule at
$\beta=2$. A \emph{product} of $p$ Householders
$R_t = \prod_{j \le p}(I - 2 u_t^{(j)} u_t^{(j)\tr})$ gives a richer orthogonal
transform, and the WY/UT-transform kernel machinery already represents products of
rank-one reflectors, so it stays chunkable and linear-time ($p = 2\text{--}4$).
Data-dependent \textbf{Givens} / block-diagonal $2\times2$ rotations are a cheaper
$O(d_k)$ alternative and a bridge to Mamba-3's complex rotations. Here $D_t$ is the
\emph{only} contraction: reorganization is norm-preserving, forgetting explicit and
separate.

\subsection{A2 --- hybrid}
A0-capped small erase (associative overwrite) $+$ A1 orthogonal reorganization keeps
the delta rule's ``remove old value at $k$ before writing'' while adding
norm-preserving structure. \emph{Honest tension:} replacing erase-by-subtraction with
rotation (A1) weakens associative overwrite; A2 mitigates, and A0 is the safe first
experiment.


% =============================================================================
\section{Direction B: Multi-Stream Hyper-Memory}
\label{sec:dirB}

We replace the single state $S$ with $n$ streams $S^{(1)},\dots,S^{(n)}$ in a shared
$d_k\times d_v$ space that exchange information via a doubly-stochastic mixing ---
hyper-connected memory lanes. Per step, each stream runs its own
(A-constrained) update, then streams mix:
\begin{align}
  \widetilde{S}_t^{(i)} &= \Tcarry_t^{(i)} S_{t-1}^{(i)} + k_t^{(i)} z_t^{(i)\tr},
  \label{eq:b-stream}\\
  S_t^{(i)} &= \textstyle\sum_j H[i,j]\, \widetilde{S}_t^{(j)},
  \qquad H \text{ doubly stochastic}.
  \label{eq:b-mix}
\end{align}
$H \in \R^{n\times n}$ is produced by Sinkhorn (Birkhoff) or Newton--Schulz
(orthostochastic).

\paragraph{Efficiency: mix at chunk boundaries.} Applying $H$ per token to full
states would break the chunkwise WY structure. Instead we mix only at chunk
boundaries: within a chunk the $n$ streams evolve independently in parallel (the
existing per-head kernel is untouched), and at each boundary the $n$ carried states
are mixed,
$S_{[c+1]}^{(i)} = \sum_j H[i,j]\,(\text{chunk-update of stream }j\text{ from }
S_{[c]}^{(j)})$. Boundaries number $O(L/C)$, so the $n\times n$ mix is negligible and
linear time is preserved.

\paragraph{Why doubly-stochastic here.} Forgetting is handled \emph{inside} each
stream's $\Tcarry_t$; cross-stream mixing should only route information without
amplifying any stream or letting one dominate --- i.e.\ mass-preserving, doubly
stochastic. By Birkhoff--von Neumann, $H = \sum_p \lambda_p P_p$ is a convex
combination of permutations, so $\|H\|_2 \le 1$. This is mHC's original justification,
now placed on the memory axis.

\paragraph{What differs across streams.} To make mixing meaningful, streams must
differ. Our primary instantiation is \textbf{B-timescale}: shared projections but
different decay ranges per stream (fast $\to$ slow), so mixing lets fast and slow
memory trade content each chunk --- a multi-timescale associative memory aimed
directly at long context. Alternatives: \textbf{B-heads} (per-stream Q/K/V/gate
projections, needing a shared readout space) and \textbf{B-budget} (per-stream
Householder budget $p$).

\paragraph{Static vs.\ dynamic $H$, readout, cost.} We start with a static per-layer
$H$ (cheapest, most stable), initialized near identity via
$H = (1-\alpha)I + \alpha\,\mathrm{Sinkhorn}(\text{logits})$; a data-dependent $H_t$
re-projected each chunk is an optional extension. Readout sums (or concatenates) the
per-stream outputs $o_t = \sum_i (S_t^{(i)})^{\tr} q_t^{(i)}$. Cost is $n\times$ state
memory (small $n\!=\!2\text{--}4$), $n^2$ logits per layer, a cheap Sinkhorn, and a
negligible boundary mix. Relative to prior art --- multi-head linear attention (no
state mixing), Mamba-3 MIMO (multi-rank, not doubly-stochastic cross-state mixing),
and hyper-connections (depth) --- B is hyper-connections along the \emph{memory} axis
of linear attention.


% =============================================================================
\section{Composition: Provably Non-Expansive Memory}
\label{sec:compose}

The two directions compose. If each per-stream carry $\Tcarry_t^{(i)}$ is
non-expansive (Direction A) and $H$ is doubly stochastic
($\|H\|_2 \le 1$, Direction B), the whole multi-stream memory operator is
non-expansive, giving stable long-context memory \emph{by construction} while
retaining GDN-2's decoupled expressiveness inside each stream:
\begin{align}
  \text{per stream:}\quad & \widetilde{S}^{(i)}
    = D^{(i)} R^{(i)} S^{(i)}_{\text{prev}} + k z^{(i)\tr}
    && \text{(A: non-expansive / orthogonal)}\\
  \text{cross stream:}\quad & S^{(i)}
    = \textstyle\sum_j H[i,j]\, \widetilde{S}^{(j)}
    && \text{(B: mass-preserving mix)} .
\end{align}
\begin{proposition}[Non-expansive multi-stream memory]
\label{prop:compose}
Stack the streams as $\mathbf S = (S^{(1)},\dots,S^{(n)})$ and equip the product space
with the norm $\|\mathbf S\|^2 = \sum_i \|S^{(i)}\|_F^2$. Suppose every per-stream carry
is non-expansive, $\smax(\Tcarry^{(i)}) \le 1$, and $H$ is doubly stochastic. Then the
homogeneous one-step propagation
$\mathbf S \mapsto \mathcal{M}\,\mathbf S$, where
$(\mathcal M \mathbf S)^{(i)} = \sum_j H[i,j]\,\Tcarry^{(j)} S^{(j)}$, is non-expansive:
$\|\mathcal M \mathbf S\| \le \|\mathbf S\|$.
\end{proposition}
\begin{proof}
Write $\mathcal M = \mathcal H \circ \mathcal T$, where
$\mathcal T = \mathrm{blockdiag}(\Tcarry^{(1)},\dots,\Tcarry^{(n)})$ applies each carry
within its stream and $\mathcal H$ applies the mixing $H$ across streams. For $\mathcal T$,
$\|\mathcal T\mathbf S\|^2 = \sum_i \|\Tcarry^{(i)} S^{(i)}\|_F^2
\le \sum_i \smax(\Tcarry^{(i)})^2 \|S^{(i)}\|_F^2 \le \sum_i \|S^{(i)}\|_F^2 = \|\mathbf S\|^2$,
so $\mathcal T$ is non-expansive. Viewing each $S^{(i)}$ as a vector in $\R^{d_k d_v}$,
$\mathcal H$ is the linear map $H \otimes I_{d_k d_v}$ on $\R^{\,n\, d_k d_v}$, whose
spectral norm is $\|H\otimes I\|_2 = \|H\|_2\,\|I\|_2 = \|H\|_2$. By
Birkhoff--von Neumann, a doubly-stochastic $H$ is a convex combination of permutation
matrices, so $\|H\|_2 \le 1$ and $\mathcal H$ is non-expansive. A composition of
non-expansive maps is non-expansive, hence
$\|\mathcal M\mathbf S\| \le \|\mathcal T\mathbf S\| \le \|\mathbf S\|$.
\end{proof}
Consequently old memory can only stay or shrink as it is carried and re-mixed across
chunks, no matter how many streams or how long the context --- stability by
construction, with GDN-2's decoupled expressiveness preserved inside each stream.


% =============================================================================
\section{Experimental Program}
\label{sec:exp}

We deliberately validate the premise \emph{before} kernel work.

\paragraph{1. Confirm the premise.} In a small trained GDN-2 (recurrent-only,
$\alpha\to1$ regime), instrument the distribution of $\smax(\Tcarry_t)$ and the
growth of $\|S_t\|$ and $\mathrm{cond}(S_t)$ against sequence position. \emph{Premise
risk:} decay and step-to-step decorrelation of $k_t$ may absorb the per-step
$\smax\approx1.1$; if so, the fix is unnecessary and the paper must say so. This
measurement is the gate for everything downstream.

\paragraph{2. Reference implementation first.} Implement A0/A1/B in a plain
(non-Triton) recurrence for correctness and small-scale signal before touching
kernels.

\paragraph{3. Ablation matrix.} Baseline GDN-2 vs.\ $+$A0 vs.\ $+$A1 vs.\ $+$A2 vs.\
$+$B vs.\ $+$A$+$B, evaluated on WikiText/LAMBADA perplexity, \textbf{MK-NIAH}
(GDN-2's headline multi-key retrieval), and state-norm stability. Hypotheses: A
reduces state-norm blow-up without hurting perplexity; B helps multi-key retrieval by
giving competing associations separate lanes.

\paragraph{4. Testbed.} This repository is the home for the clean reference and toy
validation; scaled experiments target a GDN / linear-attention training branch, since
softmax-attention GPT testbeds do not exercise the recurrent memory.


% =============================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Hyper-connections and manifold constraints.} Hyper-connections
\citep{zhu2024hyperconnections} and Frac-connections \citep{zhu2025fracconnections}
generalize residual streams;
mHC \citep{deepseek2025mhc} constrains the residual carry to the Birkhoff or
orthostochastic manifold. We transfer this principle from the depth axis to the time
axis of linear-attention memory.

\paragraph{Delta-rule linear attention.} DeltaNet \citep{schlag2021linear} and its
chunkwise parallelization \citep{yang2024parallelizing}, Gated DeltaNet
\citep{yang2024gdn}, KDA, and GDN-2 \citep{hatamizadeh2026gdn2} form the ladder of
Table~\ref{tab:ladder}. Our theorem characterizes precisely where step-wise
non-expansiveness is lost along this ladder.

\paragraph{Spectra and state-tracking.} Negative-eigenvalue delta rules unlock
state-tracking in linear RNNs \citep{grazzi2025statetracking}; we note $b\in[0,2]$
interacts with A0's damping and flag it for measurement. Mamba-3's data-dependent
complex rotations and MIMO states are related to A1 and B respectively.


% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

We showed that GDN-2's erase/write decoupling necessarily makes its per-step memory
carry non-normal and expansive, gave a tight closed form for the spectral norm, and
recast the memory recurrence as the time-axis analog of hyper-connections. This yields
HyperDeltaNet: a damp-or-enrich dichotomy that returns the single carry to the
non-expansive manifold, a multi-stream doubly-stochastic generalization, and a
composition that is non-expansive by construction. Whether the predicted instability
is consequential at scale is an empirical question we have framed precisely; answering
it is the immediate next step.

\appendix
\section{Proof of Theorem~\ref{thm:boundary}}
\label{app:proof}

Recall $\|k\|=1$, $e = b\odot k$, $M = I - k e^{\tr}$, and set $c = k^{\tr}e$,
$p = \|e\|^2$. We compute the eigenvalues of $M^{\tr}M$.

\paragraph{Step 1: a rank-two perturbation of the identity.}
Since $\|k\|^2 = 1$,
\begin{equation*}
  M^{\tr}M = (I - e k^{\tr})(I - k e^{\tr})
           = I - e k^{\tr} - k e^{\tr} + e\,(k^{\tr}k)\,e^{\tr}
           = I + N,
  \qquad
  N := -k e^{\tr} - e k^{\tr} + e e^{\tr}.
\end{equation*}
Every column of $N$ lies in $\mathrm{span}\{k,e\}$, so $N$ vanishes on the orthogonal
complement of this (at most) two-dimensional subspace. Hence $M^{\tr}M$ has eigenvalue
$1$ with multiplicity at least $d_k - 2$, and its remaining spectrum is that of $N$
restricted to $\mathrm{span}\{k,e\}$.

\paragraph{Step 2: an orthonormal basis of $\mathrm{span}\{k,e\}$.}
Decompose $e = c\,k + f$ with $f \perp k$. Then
$q := \|f\|^2 = \|e\|^2 - c^2 = p - c^2 \ge 0$, the inequality being Cauchy--Schwarz
$c^2 = (k^{\tr}e)^2 \le \|k\|^2\|e\|^2 = p$. If $q = 0$ then $e \parallel k$ and $N$ has
rank one (treated below); otherwise take the orthonormal basis $\{k,\ \hat f\}$ with
$\hat f = f/\sqrt{q}$, so $e = c\,k + \sqrt{q}\,\hat f$.

\paragraph{Step 3: the $2\times2$ block.}
Using $k^{\tr}e = c$, $\hat f^{\tr}e = \sqrt q$, and $k^{\tr}\hat f = 0$,
\begin{align*}
  N k       &= -c\,k - e + c\,e = (c^2 - 2c)\,k + \sqrt q\,(c-1)\,\hat f, \\
  N \hat f  &= -\sqrt q\,k + \sqrt q\,e = \sqrt q\,(c-1)\,k + q\,\hat f,
\end{align*}
so in the basis $\{k,\hat f\}$ the restriction of $N$ is the symmetric block
\begin{equation*}
  B = \begin{bmatrix} c^2 - 2c & \sqrt q\,(c-1)\\[2pt] \sqrt q\,(c-1) & q \end{bmatrix},
  \qquad
  \tr B = (c^2 - 2c) + q = p - 2c,
  \qquad
  \det B = (c^2-2c)q - q(c-1)^2 = -q .
\end{equation*}

\paragraph{Step 4: eigenvalues.}
The eigenvalues of $B$ are
$\lambda_\pm = \tfrac12\big[\tr B \pm \sqrt{(\tr B)^2 - 4\det B}\,\big]$, and
\begin{equation*}
  (\tr B)^2 - 4\det B = (p-2c)^2 + 4q = p^2 - 4pc + 4c^2 + 4(p - c^2) = p\,(p + 4 - 4c),
\end{equation*}
giving $\lambda_\pm = \tfrac12\big[(p-2c) \pm \sqrt{p\,(p+4-4c)}\,\big]$. Because
$\det B = -q \le 0$, the two eigenvalues have opposite signs (or one is zero):
$\lambda_+ \ge 0 \ge \lambda_-$. The largest eigenvalue of $M^{\tr}M = I + N$ is
therefore $1 + \lambda_+$, i.e.
\begin{equation*}
  \smax(M)^2 = 1 + \lambda_+ = 1 + \tfrac12\big[(p-2c) + \sqrt{p\,(p+4-4c)}\,\big] \ge 1 .
\end{equation*}
Equality holds iff $\lambda_+ = 0$ iff $q = 0$ iff $f = 0$ iff $e \parallel k$; since
$e = b\odot k$, this is iff $b$ is a per-head scalar (the KDA/Gated-DeltaNet case). When
$q = 0$ the block $B$ collapses to $[\,c^2 - 2c\,]$ with $c^2 = p$, whose single
eigenvalue $c^2 - 2c = p - 2c \le 0$ leaves $\smax(M) = 1$, consistent with the limit.
\qed

\paragraph{Worked example.}
For $k = (\tfrac1{\sqrt2},\tfrac1{\sqrt2})$ and $b = (1,0)$, we have
$e = (\tfrac1{\sqrt2}, 0)$, hence $c = \tfrac12$, $p = \tfrac12$, $q = p - c^2 = \tfrac14$.
Then $\lambda_+ = \tfrac12\big[(\tfrac12 - 1) + \sqrt{\tfrac12(\tfrac12 + 4 - 2)}\big]
= \tfrac12\big[-\tfrac12 + \sqrt{1.25}\big] \approx 0.309$, so
$\smax(M) = \sqrt{1.309} \approx 1.144$, matching~\eqref{eq:nonnormal-ex}.

\bibliography{main}
\bibliographystyle{plainnat}


\end{document}
