\RequirePackage[svgnames,table]{xcolor}

\documentclass[11pt,letterpaper,logo]{yalearxiv}

\input{preamble}
\input{command}


\newtcolorbox{simpleElegantQuote}{
    colback=AliceBlue!50!White,
    colframe=RoyalBlue!75!Black,
    boxrule=0.5pt,
    arc=2mm,
    boxsep=4pt,
    left=10pt, right=10pt,
    top=8pt, bottom=8pt,
    fontupper=\itshape,
}

\title{HyperDeltaNet: Manifold-Constrained Memory for Linear Attention}

\runningtitle{HyperDeltaNet}

\usepackage{fontawesome5}
\definecolor{myorange}{RGB}{250,99,57}
\newcommand{\orgfire}{\textcolor{myorange}{\faFire*}}

\definecolor{yaleblue}{RGB}{0,58,112}
\newcommand{\yale}{%
  \textsuperscript{%
    {\usefont{T1}{pbk}{m}{n}\textcolor{yaleblue}{\textbf{Y}}}%
  }%
}

\author{Ngoc Bui\yale, Rex Ying\yale\\
\yale Department of Computer Science, Yale University\\
\faGithub~\textbf{Source Code:} \href{https://github.com/ngocbh/hyper-delta-net}{\texttt{https://github.com/ngocbh/hyper-delta-net}}
}

\hypersetup{colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black,
            urlcolor=blue!50!black}

% --- draft-local notation ---
\newcommand{\smax}{\sigma_{\max}}
\newcommand{\Tcarry}{T}                        % time-axis carry operator T_t
\newcommand{\Hres}{H_{\mathrm{res}}}
\newcommand{\Hpre}{H_{\mathrm{pre}}}
\newcommand{\Hpost}{H_{\mathrm{post}}}
\newcommand{\Sbb}{\boldsymbol{\mathcal{S}}}    % stacked multi-stream state
\newcommand{\Acal}{\mathcal{A}}                % stacked one-step operator
\newcommand{\Fdelta}{\mathsf{F}}               % delta-rule residual block


\begin{document}


\begin{abstract}
\vspace{-1mm}
{\centering\section*{Abstract}}
Linear-attention models compress an unbounded key--value cache into a fixed-size
recurrent state, so the central design question is how to \emph{edit} that compressed
memory without scrambling stored associations. We begin from a simple observation: the
delta rule is a \emph{residual block along time}, $S_t = S_{t-1} + \Fdelta(S_{t-1})$,
which makes linear-attention memory the time-axis analog of a hyper-connection --- the
erase/read and the write play the roles of the pre- and post-connection maps, and the
step-to-step carry plays the role of the residual mixing matrix. This viewpoint yields a
single \emph{scale-free} result. For any rank-one memory edit $M = I - u w^{\tr}$ we give
a closed form for its largest singular value and show $\smax(M) \ge 1$ always, with
equality \emph{iff} $u \parallel w$. As a \emph{within-token} corollary this recovers
exactly the non-normal amplification that Gated DeltaNet-2's (GDN-2) erase/write
decoupling introduces; as a \emph{cross-stream} corollary the identical statement governs
a new multi-stream memory. The impossibility forces two remedies at both scales ---
\emph{damp} the edit onto the non-expansive set, or \emph{enrich} it with orthogonal
structure --- which we develop as HyperDeltaNet. We then generalize the single state into
$n$ doubly-stochastically mixed streams, prove the whole memory operator is non-expansive
by construction, and characterize exactly when the streams remain non-degenerate. We lay
out an honest experimental program to test whether the predicted instability is
empirically consequential and whether the fix improves long-context retrieval.
\end{abstract}


\maketitle


% =============================================================================
\section{Introduction}
\label{sec:intro}

Linear attention replaces the growing key--value cache of softmax attention with a
fixed-size recurrent state $S_t \in \R^{d_k \times d_v}$, trading unbounded memory for
constant-time, constant-memory inference \citep{katharopoulos2020transformers}. The price
is that every token must be \emph{written into} a compressed store that cannot grow: the
model must continually decide what to overwrite, what to preserve, and what to forget. The
delta-rule family \citep{schlag2021linear,yang2024parallelizing} made this an explicit,
associative edit --- remove the old value bound to a key before writing the new one --- and
successive refinements (Gated DeltaNet \citep{yang2024gdn}, Kimi Delta Attention) added
gating and channel-wise decay for finer control. Gated DeltaNet-2 (GDN-2)
\citep{hatamizadeh2026gdn2} pushes this line furthest by \emph{decoupling} the erase and
write decisions that all prior delta-rule models had tied to a single scalar gate: a
channel-wise \emph{erase} gate $b_t$ selects which key-side coordinates to remove, and a
channel-wise \emph{write} gate $w_t$ selects which value-side coordinates to commit. This
decoupling is GDN-2's headline contribution and delivers its strongest results on
interference-heavy multi-key retrieval.

\paragraph{Memory is a residual highway along time.} Our starting point is that the delta
rule is \emph{literally a residual block}, but with the block index running over time
rather than depth:
\begin{equation}
  S_t \;=\; S_{t-1} \;+\; \Fdelta(S_{t-1}),
  \qquad
  \Fdelta(S) \;=\; \beta_t\, k_t\,(v_t - S^{\tr} k_t)^{\tr}.
  \label{eq:delta-res}
\end{equation}
The skip connection is $S_{t-1}$; the ``layer'' $\Fdelta$ reads the state through $k_t$,
computes the prediction error $v_t - S_{t-1}^{\tr}k_t$, and writes the correction back
along $k_t$. This is exactly the structure of a hyper-connection
\citep{zhu2024hyperconnections}, whose manifold-constrained variant mHC
\citep{deepseek2025mhc} keeps the residual highway norm-controlled across \emph{depth}. We
ask whether the same discipline stabilizes the memory highway across \emph{time} --- and
find that GDN-2's decoupling is precisely what makes it necessary.

Reading~\eqref{eq:delta-res} through the hyper-connection lens exposes the full
correspondence, not just the carry (Table~\ref{tab:halves}): the read direction is the
analog of the pre-connection $\Hpre$, the write direction of the post-connection
$\Hpost$, and the step-to-step carry of the residual mixing $\Hres$. GDN-2's erase/write
decoupling is exactly $\Hpre \neq \Hpost$; what it never constrains is the carry.

\paragraph{One theorem, two scales.} The technical core is a single statement about
rank-one memory edits $M = I - u w^{\tr}$ (Theorem~\ref{thm:master}): a closed form for
$\smax(M)$, which is $\ge 1$ always and equal to $1$ only when $u \parallel w$. Applied
\emph{within} a token it characterizes GDN-2's non-normal amplification
(Corollary~\ref{cor:within}); applied \emph{across} streams it governs a multi-stream
memory (Corollary~\ref{cor:cross}). The two uses are the same algebra at different scales,
and the same fix --- damp or enrich --- works at both.

\paragraph{Contributions.}
\begin{itemize}[leftmargin=1.5em,itemsep=1pt,topsep=2pt]
  \item \textbf{Memory as a time-axis hyper-connection (\S\ref{sec:view}).} We recast the
  delta-rule recurrence as a residual block and map the full $\Hpre/\Hres/\Hpost$ triple
  onto read/carry/write, locating GDN-2's decoupling precisely within it.
  \item \textbf{A scale-free impossibility theorem (\S\ref{sec:view}).} A closed form for
  $\smax(I - u w^{\tr})$ with $\smax \ge 1$, equality iff $u \parallel w$. Its
  within-token corollary recovers GDN-2's expansive carry; its cross-stream corollary
  governs multi-stream memory.
  \item \textbf{Direction A --- constraining the carry (\S\ref{sec:dirA}).} The theorem
  forces two escape routes: \emph{damp} onto the non-expansive set (A0, closed-form
  drop-in) or \emph{enrich} with orthogonal structure (A1, Householder/Givens); both apply
  verbatim at either scale.
  \item \textbf{Direction B --- multi-stream hyper-memory (\S\ref{sec:dirB}).} A faithful
  time-axis hyper-connection over $n$ streams, with a taxonomy (shared-block, channel-gate,
  per-stream-block) that cleanly separates multi-timescale memory from multi-key retrieval,
  and a non-degeneracy condition on the streams.
  \item \textbf{Composition (\S\ref{sec:compose}).} Constrained edits and mass-preserving
  mixing compose into a memory operator that is non-expansive by construction.
  \item \textbf{An honest experimental program (\S\ref{sec:exp}).} The measurements that
  would confirm or refute that the predicted amplification is consequential at scale,
  before any kernel work.
\end{itemize}


% =============================================================================
\section{Background}
\label{sec:bg}

\subsection{Two residuals}
\label{sec:two-residuals}

\paragraph{The depth axis: hyper-connections and mHC.} With $s$ residual streams, a
hyper-connection layer updates
\begin{equation}
  x_{l+1} \;=\; \Hres\, x_l \;+\; \Hpost^{\tr}\, F\!\big(\Hpre\, x_l,\, W_l\big),
  \label{eq:hc}
\end{equation}
where $\Hres \in \R^{s\times s}$ mixes streams and $\Hpre,\Hpost$ are \emph{separate}
read/write maps. mHC \citep{deepseek2025mhc} constrains $\Hres$ to a norm-controlled
manifold --- the Birkhoff polytope (doubly stochastic, via Sinkhorn) or the orthostochastic
set (squared entries of an orthogonal matrix, via Newton--Schulz) --- so the deep residual
highway stays norm-controlled across depth. Frac-connections
\citep{zhu2025fracconnections} are a related fractional generalization.

\paragraph{The time axis: the delta-rule ladder.} Linear-attention memories share the
recurrence $S_t = \Tcarry_t\, S_{t-1} + (\text{write})$, output $o_t = S_t^{\tr} q_t$.
Table~\ref{tab:ladder} places the main variants on a ladder by their carry operator
$\Tcarry_t$. Reading down, the carry grows from the identity to a scalar decay, to a
symmetric rank-one projector (the delta rule), to gated and channel-wise-decayed versions,
and finally to GDN-2's \emph{non-symmetric} carry.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Model & State update $S_t=\dots$ & Carry on $S_{t-1}$ \\
\midrule
Linear attn & $S_{t-1} + k_t v_t^{\tr}$ & $I$ \\
Mamba-2 & $\alpha_t S_{t-1} + k_t v_t^{\tr}$ & $\alpha_t I$ \\
DeltaNet & $(I - \beta_t k_t k_t^{\tr})S_{t-1} + \beta_t k_t v_t^{\tr}$ & $I - \beta_t k_t k_t^{\tr}$ (\textbf{sym.}) \\
Gated DeltaNet & $\alpha_t(I - \beta_t k_t k_t^{\tr})S_{t-1} + \beta_t k_t v_t^{\tr}$ & $\alpha_t(I - \beta_t k_t k_t^{\tr})$ \\
KDA & $(I - \beta_t k_t k_t^{\tr})D_t S_{t-1} + \beta_t k_t v_t^{\tr}$ & $(I - \beta_t k_t k_t^{\tr})D_t$ \\
\textbf{GDN-2} & $(I - k_t e_t^{\tr})D_t S_{t-1} + k_t z_t^{\tr}$ & $(I - k_t e_t^{\tr})D_t$ (\textbf{non-sym.}) \\
\bottomrule
\end{tabular}
\caption{The delta-rule ladder. $D_t=\diag(\alpha_t)$; for GDN-2, $e_t = b_t \odot k_t$
(channel-wise erase) and $z_t = w_t \odot v_t$ (channel-wise write). GDN-2 recovers KDA
when $b_t = w_t = \beta_t \Onef$, and Gated DeltaNet when the decay is also scalar.}
\label{tab:ladder}
\end{table}

\paragraph{The residual form.} Expanding~\eqref{eq:delta-res} gives
$S_t = (I - \beta_t k_t k_t^{\tr})S_{t-1} + \beta_t k_t v_t^{\tr}$: the carry
$I - \beta k k^{\tr}$ is the erase, hidden inside the residual block's dependence on
$S_{t-1}$. The same rearrangement puts GDN-2 in residual form,
\begin{equation}
  S_t \;=\; D_t S_{t-1} \;+\; k_t\big(z_t - S_{t-1}^{\tr}(D_t e_t)\big)^{\tr},
  \label{eq:gdn2-res}
\end{equation}
exposing the read direction $D_t e_t = D_t(b_t \odot k_t)$ and the write direction
$z_t = w_t \odot v_t$ as \emph{distinct} --- the $\Hpre \neq \Hpost$ decoupling. The carry
that governs stability is nonetheless the erase factor $I - k_t e_t^{\tr}$, analyzed next.

\subsection{Non-expansive vs.\ expansive carries}
\label{sec:sigmax}

A linear map $M$ is \emph{non-expansive} if $\|Mx\| \le \|x\|$ for all $x$, i.e.\
$\smax(M) = \|M\|_2 \le 1$; \emph{expansive} if some direction is stretched
($\smax(M) > 1$). Because the recurrence multiplies the state by $\Tcarry_t$ every step,
old memory passes through the product $\Tcarry_L \cdots \Tcarry_1$; a non-expansive carry
lets old memory only stay or shrink (legal forgetting), while an expansive carry can let it
grow step after step and swamp new writes.

Eigenvalues govern the \emph{asymptotic} fate; singular values govern a \emph{single} step.
A \emph{non-normal} matrix (non-orthogonal eigenvectors) can have all eigenvalues $\le 1$
yet still stretch transiently:
\begin{equation}
  k=(\tfrac1{\sqrt2},\tfrac1{\sqrt2}),\ \ b=(1,0)
  \ \Rightarrow\ e = b\odot k = (\tfrac1{\sqrt2},0),\ \
  M = I - k e^{\tr} = \begin{bmatrix} 0.5 & 0\\ -0.5 & 1\end{bmatrix},\ \
  \text{eig}=\{0.5,1\},\ \smax\approx 1.14 .
  \label{eq:nonnormal-ex}
\end{equation}
Symmetric matrices are normal, so eigenvalues equal singular values --- for DeltaNet/GDN/KDA
``eigenvalues $\le 1$'' already implies non-expansive. GDN-2's decoupling makes the erase
non-symmetric and opens the gap.


% =============================================================================
\section{Memory as a Time-Axis Hyper-Connection}
\label{sec:view}

\subsection{The correspondence}
Define the time-axis carry $\Tcarry_t = (I - k_t e_t^{\tr}) D_t$; then
$S_t = \Tcarry_t S_{t-1} + k_t z_t^{\tr}$ is structurally identical to the
hyper-connection update~\eqref{eq:hc}. Combined with the residual
form~\eqref{eq:gdn2-res}, this gives the correspondence that organizes the paper:
\begin{center}
\emph{$\Tcarry_t$ is the time-axis analog of $\Hres$; the read direction $D_t e_t$ and the
write direction $z_t$ are the analogs of $\Hpre$ and $\Hpost$.}
\end{center}
Stability of the recurrence is governed by the cumulative product
$\Tcarry_t \cdots \Tcarry_1$ --- exactly what mHC controls for depth. Each prior model
holds only half of mHC's discipline (Table~\ref{tab:halves}).

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
 & Constrained carry ($\Hres$-like) & Separated read/write ($\Hpre \neq \Hpost$)\\
\midrule
DeltaNet/KDA & \checkmark\ non-expansive & $\times$\ tied to $k$\\
GDN-2        & $\times$\ $\smax > 1$      & \checkmark\ decoupled\\
\textbf{mHC / HyperDeltaNet} & \checkmark & \checkmark\\
\bottomrule
\end{tabular}
\caption{Each prior model holds half of mHC's discipline. HyperDeltaNet restores both on
the time axis: a constrained carry \emph{and} decoupled read/write.}
\label{tab:halves}
\end{table}

\subsection{A scale-free theorem for rank-one edits}

Every memory edit in the delta-rule family, and every cross-stream mix we introduce later,
is a rank-one update of the identity. The following closed form is the technical spine of
the paper; it is stated once, in full generality, and used at two scales.

\begin{theorem}[Rank-one memory edits]
\label{thm:master}
Let $u, w \in \R^N$ and $M = I - u w^{\tr}$. Write $a = \|u\|^2$, $b = \|w\|^2$, and
$r = u^{\tr} w$. Then
\begin{equation}
  \smax(M)^2 \;=\; 1 + \tfrac12\!\left[(ab - 2r) + \sqrt{ab\,(ab - 4r + 4)}\,\right]
  \;\ge\; 1 ,
  \label{eq:master}
\end{equation}
with equality (i.e.\ $M$ non-expansive, $\smax(M)=1$) \emph{if and only if}
$u \parallel w$ and $u^{\tr} w \in [0,2]$.
\end{theorem}

The proof (Appendix~\ref{app:master}) reduces $M^{\tr}M = I + N$ to a $2\times2$ block on
$\mathrm{span}\{u,w\}$ with trace $ab - 2r$ and determinant $r^2 - ab \le 0$
(Cauchy--Schwarz); the non-positive determinant forces the larger eigenvalue of $N$ to be
$\ge 0$, hence $\smax \ge 1$, with equality only in the aligned case within the gain window.

\begin{corollary}[Within-token: GDN-2's carry is expansive]
\label{cor:within}
For GDN-2's erase factor $M = I - k e^{\tr}$ with $\|k\|=1$ and $e = b \odot k$
(so $a=1$), set $c = k^{\tr}e = \sum_i b_i k_i^2$ and $p = \|e\|^2 = \sum_i b_i^2 k_i^2$.
Then
\begin{equation}
  \smax(M)^2 = 1 + \tfrac12\!\left[(p - 2c) + \sqrt{p\,(p + 4 - 4c)}\,\right] \ge 1 ,
  \label{eq:within}
\end{equation}
with equality iff $e \parallel k$ iff $b$ is a per-head scalar (the KDA/Gated-DeltaNet
case). The worked example~\eqref{eq:nonnormal-ex} gives $c = p = \tfrac12$ and
$\smax \approx 1.14$.
\end{corollary}

\begin{corollary}[Who is safe]
\label{cor:safe}
DeltaNet, Gated DeltaNet, and KDA are step-wise non-expansive: their erase factor
$I - \beta k k^{\tr}$ ($\|k\|=1$, $\beta \in [0,2]$) is symmetric with eigenvalues
$\{1^{(d_k-1)}, 1-\beta\}$, so $\smax = 1$; KDA remains safe under $D_t$ as a product of
two individually non-expansive factors. GDN-2's carry, by Corollary~\ref{cor:within}, has
$\smax > 1$ whenever $b$ is genuinely channel-wise.
\end{corollary}

\paragraph{Why this matters where GDN-2 wins.} The decay $D_t$ can mask the extra gain only
when $\alpha$ is small; but the long-memory regime $\alpha \to 1$ --- where GDN-2 reports
its long-context RULER / MK-NIAH gains --- is exactly where masking vanishes. GDN-2's core
innovation ($b$ channel-wise $\Rightarrow e \not\parallel k$) \emph{necessarily} sacrifices
step-wise non-expansiveness, and the model never constrains it.

\subsection{Eq.~5 as the on-manifold anchor}
The plain delta rule's carry $I - \beta k k^{\tr}$ traces a one-parameter path from the
identity ($\beta=0$) through the orthogonal projector ($\beta=1$) to a Householder
reflection ($\beta=2$) --- inside the norm-controlled orthogonal family that mHC's
orthostochastic/Newton--Schulz variant is built around. It is also a stable optimizer step:
the delta rule is one gradient step on $\tfrac12\|S^{\tr}k - v\|^2$ (Hessian $kk^{\tr}$), so
the carry is $I - \beta\cdot\text{Hessian}$, stable for $\beta \le 2/\lambda_{\max} = 2$ ---
the same window that appears in Theorem~\ref{thm:master}'s equality condition
$u^{\tr}w \in [0,2]$. GDN-2 stepped off this manifold to gain expressiveness; the rest of
the paper puts each carry back on the manifold correct for its axis.


% =============================================================================
\section{Direction A: Constraining the Single Carry}
\label{sec:dirA}

\paragraph{The dichotomy.} Theorem~\ref{thm:master} says a rank-one edit is non-expansive
\emph{iff} $u \parallel w$ (within the gain window): one cannot have \emph{decoupled}
($u \not\parallel w$) \textbf{and} \emph{rank-one} \textbf{and} \emph{non-expansive} at
once. Relaxing exactly one of the latter two yields the two families below; both apply
verbatim at the within-token scale (here) and the cross-stream scale
(\S\ref{sec:compose}). Which manifold is correct depends on the axis
(Table~\ref{tab:manifolds}).

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
Axis & Operator & Correct manifold & Why \\
\midrule
Within a state (time) & carry $\Tcarry_t$ & non-expansive / orthogonal &
forgetting must stay legal (contraction ok, not mass-preserving) \\
Across states (streams) & mixing $H$ & doubly-stochastic / orthostochastic &
forgetting handled per-stream; mixing should preserve mass \\
\bottomrule
\end{tabular}
\caption{The right mHC manifold on each axis. Birkhoff (doubly stochastic) is
\emph{wrong inside} one state (it forbids forgetting) but \emph{right across} states ---
making Directions A and B complementary, not redundant.}
\label{tab:manifolds}
\end{table}

\subsection{A0 --- non-expansive projection (drop-in)}
Per token, compute the two scalars $c = k_t^{\tr} e_t$ and $p = \|e_t\|^2$ ($O(d_k)$),
form $\smax(M_t)$ in closed form via~\eqref{eq:within}, and rescale:
\begin{equation}
  \widetilde{\Tcarry}_t = \frac{M_t D_t}{\max\!\big(1,\, \smax(M_t)\big)},
  \qquad M_t = I - k_t e_t^{\tr}.
  \label{eq:a0}
\end{equation}
This guarantees $\|\widetilde{\Tcarry}_t\| \le 1$ every step, kills the non-normal
blow-up, and keeps the erase/write decoupling. The cost is two dot products and a scalar
divide, folded into the decay --- nothing else about GDN-2 changes. It is the faithful
analog of Sinkhorn-projecting $\Hres$, adapted to the rank-one$+$diagonal structure and to
the \emph{non-expansive} set. Variants: \textbf{A0-whole} (divide the whole transition,
adding adaptive decay), \textbf{A0-corr} (damp only the correction $k e^{\tr}$, softer),
\textbf{A0-soft} (a smooth auxiliary penalty $\lambda\,\mathrm{relu}(\smax-1)^2$ with no
forward change), and \textbf{A0-cap-b} (rescale $b_t$ toward the aligned direction just
enough to hit $\smax \le 1+\eps$).

\subsection{A1 --- orthogonal by construction}
Factor the transition into \emph{reorganize} then \emph{forget}:
\begin{equation}
  S_t = D_t\, R_t\, S_{t-1} + k_t z_t^{\tr}, \qquad R_t \text{ orthogonal}.
  \label{eq:a1}
\end{equation}
The single Householder $R_t = I - 2 k_t k_t^{\tr}$ is exactly the delta rule at $\beta=2$.
A \emph{product} of $p$ Householders
$R_t = \prod_{j \le p}(I - 2 u_t^{(j)} u_t^{(j)\tr})$ gives a richer orthogonal transform,
and the WY/UT-transform kernel machinery already represents products of rank-one
reflectors, so it stays chunkable and linear-time ($p = 2\text{--}4$). Data-dependent
\textbf{Givens} / block-diagonal $2\times2$ rotations are a cheaper $O(d_k)$ alternative and
a bridge to Mamba-3's complex rotations. Here $D_t$ is the \emph{only} contraction:
reorganization is norm-preserving, forgetting explicit and separate.

\subsection{A2 --- hybrid}
A0-capped small erase (associative overwrite) $+$ A1 orthogonal reorganization keeps the
delta rule's ``remove old value at $k$ before writing'' while adding norm-preserving
structure. \emph{Honest tension:} replacing erase-by-subtraction with rotation (A1) weakens
associative overwrite; A2 mitigates, and A0 is the safe first experiment.


% =============================================================================
\section{Direction B: Multi-Stream Hyper-Memory}
\label{sec:dirB}

We now take the residual form~\eqref{eq:delta-res} at face value and extend it the way
hyper-connections extend residuals: from one stream to $n$. Let
$\Sbb_{t} = (S_t^{(1)},\dots,S_t^{(n)})$ be $n$ states in a shared $d_k \times d_v$ space.
Applying the hyper-connection template~\eqref{eq:hc} term-by-term:
\begin{align}
  \text{read } (\Hpre): \quad
    & \bar S_{t-1} = \textstyle\sum_i p_i\, S_{t-1}^{(i)},
    \label{eq:b-read}\\
  \text{block } (\Fdelta): \quad
    & G_t = \beta_t\, k_t\,(v_t - \bar S_{t-1}^{\tr} k_t)^{\tr},
    \label{eq:b-block}\\
  \text{carry}+\text{write } (\Hres,\Hpost): \quad
    & S_t^{(i)} = \textstyle\sum_j A_{ij}\, S_{t-1}^{(j)} + q_i\, G_t ,
    \label{eq:b-update}
\end{align}
with read/write stream profiles $p, q \in \R^n$ and cross-stream carry $A \in \R^{n\times n}$.
Setting $n=1$, $A=1$, $p=q=1$ recovers the plain delta rule~\eqref{eq:delta-res}; setting
$A = D_t$, $p = D_t e_t$, $q = z_t$ recovers GDN-2~\eqref{eq:gdn2-res}. To keep both
forgetting (per stream) and non-degeneracy (across streams), we factor the carry as
$A = H \cdot \diag(D^{(1)},\dots,D^{(n)})$: a per-stream decay composed with a
doubly-stochastic cross-stream mix $H$ produced by Sinkhorn (Birkhoff) or Newton--Schulz
(orthostochastic).

\subsection{A taxonomy of streams}
How the streams are allowed to differ determines what the architecture can represent
(Table~\ref{tab:streams}). A \emph{single} shared block $\Fdelta$ (one delta computation,
scattered by $q$) is the faithful and cheapest reading of hyper-connections; richer variants
recover separate associative lanes at the cost of $n$ blocks.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
Variant & Streams differ by & Semantics & Block cost \\
\midrule
\textbf{Shared-block (B-timescale)} & $(D^{(i)}, p_i, q_i)$ scalars & multi-\emph{timescale}, one store & $1$ delta \\
Channel-gate & $(b^{(i)}, w^{(i)})$ channel masks & shared keys, different channels & $1$ delta, $n$ gates \\
Per-stream-block (B-heads) & $(k^{(i)}, v^{(i)})$ projections & separate \emph{lanes} & $n$ deltas \\
\bottomrule
\end{tabular}
\caption{The shared-block variant is the literal time-axis hyper-connection: $n$ streams
are scalar-filtered views of \emph{one} associative store, ideal for multi-timescale
long-context memory. Per-stream-block gives competing keys their own lanes, targeting
multi-key retrieval (MK-NIAH) --- these are complementary axes, not a strict ordering.}
\label{tab:streams}
\end{table}

\subsection{Non-degeneracy of the streams}
A shared block scatters the \emph{same} correction $G_t$ to every stream, so the streams
must be distinguished by their carries and profiles or they collapse.

\begin{proposition}[Non-degeneracy]
\label{prop:collapse}
Scalarize the per-stream decays to $\gamma_i$ and let the homogeneous cross-stream map be
$M = H\Gamma$, $\Gamma = \diag(\gamma_i)$. If all $\gamma_i$ are equal and $p,q$ are
uniform, then $S_t^{(i)} = S_t^{(j)}$ for all $i,j,t$ (exact collapse). If the $\gamma_i$
are distinct, $M$ is a primitive nonnegative matrix whose Perron vector is not proportional
to $\Onef$, and the per-stream impulse responses $[M^\tau]_{i,\cdot}\,q$ are linearly
independent. Hence the streams carry independent memory iff they differ in decay or in
$(p_i,q_i)$.
\end{proposition}

Distinct per-stream decay is therefore a \emph{necessary} design choice, which is exactly
why the shared-block variant is instantiated as B-timescale (fast$\to$slow decays). We
initialize $H = (1-\alpha)I + \alpha\,\mathrm{Sinkhorn}(\text{logits})$ near the identity so
$H$ does not average the streams together before the decays separate them.

\subsection{Efficiency and readout}
Applying $H$ per token to full states would break the chunkwise WY structure. Instead we mix
only at chunk boundaries: within a chunk the $n$ streams evolve independently in parallel
(the existing per-head kernel is untouched), and at each boundary the $n$ carried states are
mixed. Boundaries number $O(L/C)$, so the $n\times n$ mix is negligible and linear time is
preserved. Readout uses the \emph{same} $p$-weighted combined memory the block reads
(from the post-update state), $o_t = \bar S_t^{\tr} q_t = \big(\sum_i p_i\, S_t^{(i)}\big)^{\tr} q_t$,
rather than a bare per-stream sum $\sum_i (S_t^{(i)})^{\tr} q_t$. This is what makes
reduction to GDN-2 a \emph{stable} fixed point rather than merely an algebraic one: a
dormant stream ($p_i=0$) receives \emph{zero} readout gradient, so on tasks that need only
one timescale the optimizer leaves the extra streams off instead of being driven to wake
them (the bare sum gives a dormant stream a nonzero wake-up gradient $\partial \mathcal
L/\partial p_i \propto (\partial \mathcal L/\partial o_t)\cdot G_t$, and empirically dragged
$n>1$ \emph{below} $n=1$ on single-timescale recall; \S\ref{sec:exp}).
Cost is $n\times$ state memory (small $n=2\text{--}4$), $n^2$ logits per layer, and a cheap
Sinkhorn. A static per-layer $H$ is the cheapest and most stable default; a data-dependent
$H_t$ re-projected each chunk (mean-pool the chunk hidden states $\to$ MLP $\to$ $n{\times}n$
logits $\to$ log-domain Sinkhorn $\to$ near-identity blend $H=(1-\alpha)I+\alpha\,\mathrm{DS}$)
is implemented and used below.


% =============================================================================
\section{Composition: Provably Non-Expansive Memory}
\label{sec:compose}

The cross-stream mix is itself a rank-one-structured edit, so Theorem~\ref{thm:master}
returns --- one scale up. Substituting the shared read~\eqref{eq:b-read} into the
update~\eqref{eq:b-update} and dropping the input term, the homogeneous stacked operator is
\begin{equation}
  \Acal \;=\; (H\!\cdot\!\diag(D^{(i)})) \otimes I \;-\; \beta\,(q p^{\tr}) \otimes (k k^{\tr}),
  \qquad
  \Acal_0 := I - u w^{\tr},\ \ u = \sqrt{\beta}\,(q \otimes k),\ w = \sqrt{\beta}\,(p \otimes k),
  \label{eq:cross}
\end{equation}
where $\Acal_0$ isolates the cross-stream erase (set $A=I$). This is exactly the object of
Theorem~\ref{thm:master}, with $a = \beta\|q\|^2$, $b = \beta\|p\|^2$, $r = \beta\,p^{\tr}q$.

\begin{corollary}[Cross-stream: same theorem, one scale up]
\label{cor:cross}
The cross-stream erase $\Acal_0$ is non-expansive if and only if $p \parallel q$ and
$\beta\,p^{\tr}q \in [0,2]$; otherwise $\smax(\Acal_0) > 1$. The stream-level decoupling
$p \neq q$ is the exact analog of GDN-2's within-token decoupling $e \neq k$, and the same
remedies A0/A1 apply --- with the \emph{free safe default} $p = q$ (symmetric stream
read/write), which stays multi-timescale via distinct decays.
\end{corollary}

To compose the pieces cleanly, order the step as three sub-maps: the (constrained) shared
edit, then per-stream decay, then cross-stream mix.

\begin{proposition}[Non-expansive multi-stream memory]
\label{prop:compose}
Equip the product space with $\|\Sbb\|^2 = \sum_i \|S^{(i)}\|_F^2$. If the shared edit
$\Phi$ is non-expansive (Corollary~\ref{cor:cross}, i.e.\ $p \parallel q$ in the gain
window, or A0-damped), each per-stream decay has $\smax(D^{(i)}) \le 1$, and $H$ is doubly
stochastic, then the one-step map $\Sbb_t = (H \otimes I)\,(\mathcal D)\,\Phi(\Sbb_{t-1})$ is
non-expansive.
\end{proposition}
\begin{proof}
$\Phi$ is non-expansive by hypothesis. $\mathcal D = \mathrm{blockdiag}(D^{(i)})$ satisfies
$\|\mathcal D \Sbb\|^2 = \sum_i \|D^{(i)} S^{(i)}\|_F^2 \le \sum_i \|S^{(i)}\|_F^2$. Viewing
each $S^{(i)}$ as a vector, $H \otimes I$ has spectral norm $\|H\|_2 \le 1$ by
Birkhoff--von Neumann. A composition of non-expansive maps is non-expansive, so
$\|\Sbb_t\| \le \|\Sbb_{t-1}\|$.
\end{proof}

Old memory can therefore only stay or shrink as it is carried and re-mixed across chunks,
no matter how many streams or how long the context --- stability by construction, with
GDN-2's decoupled expressiveness preserved inside each stream.


% =============================================================================
\section{Experimental Program}
\label{sec:exp}

We deliberately validate the premise \emph{before} kernel work.

\paragraph{1. Confirm the premise.} In a small trained GDN-2 (recurrent-only,
$\alpha\to1$ regime), instrument the distribution of $\smax(\Tcarry_t)$ from
\eqref{eq:within} and the growth of $\|S_t\|$ and $\mathrm{cond}(S_t)$ against sequence
position. \emph{Premise risk:} decay and step-to-step decorrelation of $k_t$ may absorb the
per-step $\smax\approx1.1$; if so, the fix is unnecessary and the paper must say so. This
measurement is the gate for everything downstream, and it applies to the cross-stream carry
\eqref{eq:cross} too.

\paragraph{2. Reference implementation first.} Implement A0/A1/B in a plain (non-Triton)
recurrence for correctness and small-scale signal before touching kernels.

\paragraph{Preliminary reference-implementation findings.} The plain recurrence
(shared-block B, $n{=}4$, dynamic $H$) is in place and passes an exact $n{=}1$ reduction
test against GDN-2. Two premises hold on toy content-based recall. \emph{(i) Graceful
reduction.} With the $p$-weighted readout, on \emph{single}-timescale recall (where extra
timescales are pure overhead) $n{=}4$ matches $n{=}1$ across seeds ($0.46$ vs.\ $0.39$ mean),
whereas the naive summed readout drove $n{=}4$ \emph{below} $n{=}1$ ($0.32$ vs.\ $0.66$) by
waking dormant streams --- the fixed-point argument of \S\ref{sec:dirB} made concrete.
\emph{(ii) Multi-timescale advantage.} On a bimodal-lag recall task (slow bindings recalled
across a long interference gap, fast bindings recalled at short lag), $n{=}4$ beats $n{=}1$
on both classes at once (overall $0.41$ vs.\ $0.27$; slow $0.33$ vs.\ $0.23$; fast $0.48$
vs.\ $0.31$, mean over 3 seeds), via an emergent fast$+$slow $\gamma$ split ($n{=}1$ is stuck
at $\gamma{\approx}1$). These are small-scale signals on the reference recurrence, not scaled
results; the ablation matrix below is the real test.

\paragraph{3. Ablation matrix.} Baseline GDN-2 vs.\ $+$A0 vs.\ $+$A1 vs.\ $+$A2 vs.\
$+$B(shared) vs.\ $+$B(per-stream) vs.\ $+$A$+$B, on WikiText/LAMBADA perplexity,
\textbf{MK-NIAH} (GDN-2's headline multi-key retrieval), and state-norm stability.
Hypotheses: A reduces state-norm blow-up without hurting perplexity; shared-block B helps
long-context by adding timescales; per-stream B helps multi-key retrieval by giving
competing associations separate lanes (Table~\ref{tab:streams}).

\paragraph{4. Testbed.} This repository is the home for the clean reference and toy
validation; scaled experiments target a GDN / linear-attention training branch, since
softmax-attention GPT testbeds do not exercise the recurrent memory.


% =============================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Hyper-connections and manifold constraints.} Hyper-connections
\citep{zhu2024hyperconnections} and Frac-connections \citep{zhu2025fracconnections}
generalize residual streams; mHC \citep{deepseek2025mhc} constrains the residual carry to
the Birkhoff or orthostochastic manifold. We transfer this principle from the depth axis to
the time axis of linear-attention memory, and show the same rank-one spectral analysis
governs both the within-token carry and cross-stream mixing.

\paragraph{Delta-rule linear attention.} DeltaNet \citep{schlag2021linear} and its chunkwise
parallelization \citep{yang2024parallelizing}, Gated DeltaNet \citep{yang2024gdn}, KDA, and
GDN-2 \citep{hatamizadeh2026gdn2} form the ladder of Table~\ref{tab:ladder}.
Theorem~\ref{thm:master} characterizes precisely where step-wise non-expansiveness is lost
along it.

\paragraph{Spectra and state-tracking.} Negative-eigenvalue delta rules unlock
state-tracking in linear RNNs \citep{grazzi2025statetracking}; we note $b\in[0,2]$ interacts
with A0's damping and flag it for measurement. Mamba-3's data-dependent complex rotations
and MIMO states are related to A1 and B respectively.


% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

Reading the delta rule as a residual block along time makes linear-attention memory a
hyper-connection, and a single closed form for rank-one edits then does double duty: it
pins down the non-normal amplification that GDN-2's erase/write decoupling necessarily
introduces, and it governs the cross-stream mixing of a new multi-stream memory. The same
damp-or-enrich remedy returns each carry to the non-expansive manifold at both scales, and
per-stream constrained edits compose with doubly-stochastic mixing into a memory operator
that is non-expansive by construction. Whether the predicted instability is consequential at
scale is an empirical question we have framed precisely; answering it is the immediate next
step.


\appendix
% -----------------------------------------------------------------------------
\section{Proof of Theorem~\ref{thm:master}}
\label{app:master}

Let $M = I - u w^{\tr}$ with $a = \|u\|^2$, $b = \|w\|^2$, $r = u^{\tr}w$.

\paragraph{Step 1: a rank-two perturbation of the identity.}
\begin{equation*}
  M^{\tr}M = (I - w u^{\tr})(I - u w^{\tr})
           = I - u w^{\tr} - w u^{\tr} + w\,(u^{\tr}u)\,w^{\tr}
           = I + N,
  \quad N := -u w^{\tr} - w u^{\tr} + a\, w w^{\tr}.
\end{equation*}
Every column of $N$ lies in $\mathrm{span}\{u,w\}$, so $M^{\tr}M$ has eigenvalue $1$ with
multiplicity at least $N-2$, and its remaining spectrum is that of $N$ restricted to
$\mathrm{span}\{u,w\}$.

\paragraph{Step 2: the $2\times2$ block.}
Write $N = X C X^{\tr}$ with $X = [\,u\ \ w\,]$ and, matching the outer products
$u u^{\tr}, u w^{\tr}, w u^{\tr}, w w^{\tr}$,
\begin{equation*}
  C = \begin{bmatrix} 0 & -1 \\ -1 & a \end{bmatrix}.
\end{equation*}
The nonzero eigenvalues of $N = X C X^{\tr}$ equal those of
$C\,X^{\tr}X = C\,G$, where $G = \begin{bmatrix} a & r \\ r & b \end{bmatrix}$ is the Gram
matrix. Hence
\begin{equation*}
  \tr(CG) = ab - 2r, \qquad
  \det(CG) = \det C \cdot \det G = (-1)(ab - r^2) = r^2 - ab \le 0,
\end{equation*}
the inequality being Cauchy--Schwarz $r^2 = (u^{\tr}w)^2 \le \|u\|^2\|w\|^2 = ab$.

\paragraph{Step 3: eigenvalues.}
The eigenvalues of $CG$ are
$\lambda_\pm = \tfrac12\big[\tr(CG) \pm \sqrt{\tr(CG)^2 - 4\det(CG)}\big]$, and
\begin{equation*}
  \tr(CG)^2 - 4\det(CG) = (ab - 2r)^2 + 4(ab - r^2) = ab\,(ab - 4r + 4),
\end{equation*}
so $\lambda_\pm = \tfrac12\big[(ab - 2r) \pm \sqrt{ab\,(ab - 4r + 4)}\big]$. Because
$\det(CG) = r^2 - ab \le 0$, the two eigenvalues have opposite signs (or one is zero):
$\lambda_+ \ge 0 \ge \lambda_-$. The largest eigenvalue of $M^{\tr}M = I + N$ is therefore
$1 + \lambda_+$, giving~\eqref{eq:master}.

\paragraph{Step 4: equality.}
$\smax(M) = 1$ iff $\lambda_+ = 0$. Squaring $\sqrt{ab(ab-4r+4)} = 2r - ab$ (which requires
$2r - ab \ge 0$) gives $ab(ab-4r+4) = (2r-ab)^2$, i.e.\ $4ab = 4r^2$, so $ab = r^2$ ---
equivalently $u \parallel w$ by Cauchy--Schwarz. Combined with $2r - ab = 2r - r^2 \ge 0$
this is $r \in [0,2]$. Conversely, when $u = \alpha w$ we have $M = I - \alpha w w^{\tr}$,
symmetric with singular values $\{1, |1 - \alpha\|w\|^2|\}$ and
$r = \alpha\|w\|^2$, so $\smax = 1$ iff $r \in [0,2]$. \qed

\paragraph{Specializations.}
Corollary~\ref{cor:within} is the case $u = k$, $w = e$ with $\|k\|=1$, so $a = 1$,
$b = p$, $r = c$, and $ab - 4r + 4 = p + 4 - 4c$. Corollary~\ref{cor:cross} is the case
$u = \sqrt\beta\,(q\otimes k)$, $w = \sqrt\beta\,(p\otimes k)$ with $\|k\|=1$, so
$a = \beta\|q\|^2$, $b = \beta\|p\|^2$, $r = \beta\, p^{\tr}q$, and $ab = r^2$ iff
$\|p\|^2\|q\|^2 = (p^{\tr}q)^2$ iff $p \parallel q$.


% -----------------------------------------------------------------------------
\section{Proof of Proposition~\ref{prop:collapse}}
\label{app:collapse}

With equal decays $\gamma_i = \gamma$ and uniform $p = q = \Onef$, every stream receives the
same carry row and the same scattered correction, so by induction $S_t^{(i)}$ is independent
of $i$. For distinct $\gamma_i$, $M = H\Gamma$ is entrywise nonnegative and, for an
irreducible aperiodic $H$, primitive; Perron--Frobenius gives a simple dominant eigenvalue
with a positive right eigenvector $v$. Were $v \propto \Onef$, then
$H\Gamma\Onef = \rho\,\Onef$ would force $H(\gamma_i)_i = \rho\,\Onef$, impossible for
doubly-stochastic $H$ unless the $\gamma_i$ are equal; hence $v \not\propto \Onef$. Finally,
the stream impulse responses are rows of $M^\tau$; since $M$ has distinct eigenvalues
(generic under distinct $\gamma_i$), the family $\{M^\tau\}_\tau$ spans a space of dimension
$n$, so the rows $[M^\tau]_{i,\cdot}\,q$ are linearly independent across $i$. \qed


\bibliography{main}
\bibliographystyle{plainnat}


\end{document}
